---
title: R session info info
author: Matt Dray
date: '2018-10-13'
slug: sessioninfo
categories:
  - R
tags:
  - sessionInfo
  - devtools
  - xfun
  - utils
---
Matt Dray

<img src ="https://media.giphy.com/media/xUA7aLFXBV7OLv8O5O/giphy.gif">

# What's session info?

It's just some details about your working environment like your version of R, the locale it's runnng from and any loaded packages.

It gives a snapshot to help people reproduce what you've done or spot any anomalies. You might get asked for this if someone is diagnosing your R problem. You may also see it at the end of blog posts or reports. 

You would typically print this to the console with `sessionInfo()`, but it gives a *lot* of information by default. Other functions have been created to make the output more friendly.

# Why?

I've been using `sessionInfo` in my posts, but read about [Yihui's `session_info()` function in his `xfun` package](https://yihui.name/xfun/) and saw [Mat](https://twitter.com/mammykins_) at [machinegurning.com](https://www.machinegurning.com/rstats/bayes_thinking/) using `session_info()` from `devtools`.

I just want to see the outputs side-by-side for my own curiosity.

# Load packages

Let's grab some packages from multiple sources to see how they're represented in the output of each function.

We'll get some packages from:

* [CRAN](https://cran.r-project.org/)
* [GitHub](http://kbroman.org/pkg_primer/pages/github.html)
* [Bioconductor](https://www.bioconductor.org/install/)

```{r warnings=FALSE, error=FALSE, message=FALSE}
# From CRAN
# install.packages("package-name")
library("dplyr")
library("purrr")
library("tidyxl")

# Dev versions of packages from rOpenSci
# devtools::install_github("github-user/package")
library("rfishbase")
library("rgbif")
library("helminthR")

# Bioconductor
source("https://bioconductor.org/biocLite.R")
biocLite()
biocLite(c("GenomicFeatures", "AnnotationDbi"))
```

# The functions

## `utils`

From the [documentation for the function](https://www.rdocumentation.org/packages/utils/versions/3.5.1/topics/sessionInfo):

> Print version information about R, the OS and attached or loaded packages.

What do you get? Loads of stuff:

* R version, plaform and OS
* matrix products, BLAS and LAPACK (linear algebra bits and bobs)
* locale
* packages -- attached base packages, other attached and loaded packages via namespace (but not attached) in separate sections with their version numbers and no notation to say where the packages came from (CRAN, GitHub or Bioconductor)

```{r}
utils::sessionInfo()
```

## `devtools` package

From the [documentation for the function](https://www.rdocumentation.org/packages/devtools/versions/1.13.6/topics/session_info):

> This is `sessionInfo()` re-written from scratch to both exclude data that's rarely useful (e.g., the full collate string or base packages loaded) and include stuff you'd like to know (e.g., where a package was installed from).

What do you get? Loads of slightly more useful stuff that's organised sensibly:

* sections split into 'session info' and 'packages'
* version, system, UI, language, locale, timezone and date
* package names with version number, date and source (CRAN, Bioconductor, Github, local)

```{r}
devtools::session_info()
```

## `sessioninfo` package

From [the CRAN page for the package](https://CRAN.R-project.org/package=sessioninfo):

> It is similar to `utils::sessionInfo()`, but includes more information about packages, and where they were installed from.

What do you get? Basically the same as `utils::sessionInfo()`, but the sections are separated by lines rather than hyphens and we have a date and source (CRAN, Github and source and Bioconductor)

```{r}
sessioninfo::session_info()
```

## `xfun` package

From [Yihui Xie's Introduction to `xfun` page](https://yihui.name/xfun/):

>...it is rarely useful to print out the names of base R packages, or information about the matrix products / BLAS / LAPACK. Often times I want additional information in the session information, such as the Pandoc version when rmarkdown is used.

What do you get? Basically `utils::sessionInfo()` but no matrixproducts, BLAS or LAPACK info, plus all the packages are listed together with their version numbers. You also can't see where the package came from (CRAN, Github or Bioconductor).

```{r}
xfun::session_info()
```

# Conclusion

Yep, they basically do the same thing with slight differences. Aren't you glad you read this?