---
title: 'Quantify colour with magick'
author: 'Matt Dray'
date: '2018-11-25'
categories:
  - R
  - dataviz
tags:
  - art
  - design
  - magick
  - colour
  - color
slug: art-of-the-possible
draft: no
---

Matt Dray ([\@mattdray](https://www.twitter.com/mattdray))

![Walrus with Rainbow vomit ([via Giphy](https://giphy.com/gifs/rainbow-vomit-walrus-Dym2LTuxGJJSw?utm_source=media-link&utm_medium=landing&utm_campaign=Media%20Links&utm_term=))](https://media.giphy.com/media/Dym2LTuxGJJSw/giphy.gif)

# Colour search

Lots of online services let you sort images by colour. Some examples:

* [Google Arts & Culture](https://artsexperiments.withgoogle.com/artpalette/colors/a46053-433932-c2ad8f-b5996a-746251)
* [Designspiration](https://www.designspiration.net/search/saves/?q=%2389ebd7%20%23afff11%20%23e5ff91%20%23ff9775%20%23f38d00)
* [TinEye Multicolr](https://labs.tineye.com/multicolr/)

These services must have extracted and quantified each image's colours. There's probably a simple way of doing this in R, so let's give it a go.

By 'simple' I mean we're not going to do any hard work or consider any colour theory.[^1] And we're going to use colours like 'green' instead of 'olive drab'.[^2]

# Art of the possible

So what we're going to do is:

1. Read in an image
2. Prepare a set of 'simple colours'
3. Map the simple colours to the image
4. Quantify the colours

There's also a step 5 that I'm not covering here: create a tool where the user chooses a colour and images are returned in order of dominance for that colour. We could also write this all into a function that takes a fodler of images and returns the percentage of each colour in each image. 

# It's a kind of ImageMagick

The `magick` R package is an implementation of [ImageMagick, an open-source software suite whose emphasis is image manipulation from the command line](https://www.imagemagick.org/script/index.php). The flexibility of `magick` can be seen in the [vignette](https://cran.r-project.org/web/packages/magick/vignettes/intro.html).

`magick` was created and is maintained by [Jeroen Ooms](https://twitter.com/opencpu?lang=en), a software engineer and postdoc at [ROpenSci](https://ropensci.org/), a collective that seeks [to develop tools for open and reproducible research](https://ropensci.org/packages/).

[ROpenSci hosted a workshop](https://ropensci.org/blog/2018/10/24/commcall-nov2018/) from Ooms about working with images in R ([presentation slides](https://jeroen.github.io/munster2018/)), which caught my attention. I've used some of his code below.

# Code

First we need to load our packages.

```{r packages, message=FALSE}
# Use install.packages("package") if not yet installed
library(dplyr)  # tidy data manipulation
library(tibble)  # tidy tables
library(magick)  # image manipulation
library(knitr)  # nice html tables
```

## Read the test image

I've chosen a colourful image to use for our test case: it's a picture of a bunch of [Lego Duplo bricks](https://www.lego.com/en-gb/themes/duplo).[^3]

We'll use `image_read()` to read the JPEG as an object of class 'magick' and then `image_scale()` to reduce the image size and save some space.

Printing the image also gives us some details of format, dimensions, etc.

```{r}
# Path to the image
duplo_path <- "https://upload.wikimedia.org/wikipedia/commons/thumb/a/ac/Lego_dublo_arto_alanenpaa_2.JPG/2560px-Lego_dublo_arto_alanenpaa_2.JPG"

# Read as magick object and resize
duplo <- 
  image_read(duplo_path) %>%
  image_scale(geometry = c("x600"))

print(duplo)
```

## Prepare simple colours

We'll map a set of simple colours to the test image. This means that the colours from the test image will be replaced by the 'closest' colour from our simple set.

One way to do this is to define our simple colour set and create an image from them. In this case I'm taking just six colours.

```{r define-simple-colours}
# Generate named vector of 'simple' colours
cols_vec <- setNames(
  c("#000000", "#0000ff", "#008000", "#ff0000", "#ffffff", "#ffff00"),
  c("black", "blue", "green", "red", "white", "yellow")
)
```

Then we can plot squares of these colours, using `image_graph()` to read them as magick-class objects.[^4] My method here is inefficient and not really reproducible, but you can see the output is an image that contains our six colours.

```{r colour-squares}
# For each vector element (colour) create a square of that colour
for (i in seq_along(cols_vec)) {
  fig_name <- paste0(names(cols_vec)[i], "_square")  # create object name
  assign(
    fig_name,  # set name
    image_graph(width = 100, height = 100, res = 300)  # create magick object
  )
  par(mar = rep(0, 4))  # set plot margins
  plot.new()  # new graphics frame
  rect(0, 0, 1, 1, col = cols_vec[i], border = cols_vec[i])  # build rectangle
  assign(fig_name, magick::image_crop(get(fig_name), "50x50+10+10")) # crop
  dev.off()  # shut down plotting device
  rm(i, fig_name)  # clear up
}

# Generate names of the coloured square objects
col_square <- paste0(names(cols_vec), "_square")

# Combine magick objects (coloured squares)
simple_cols <- image_append(c(
  get(col_square[1]), get(col_square[2]), get(col_square[3]),
  get(col_square[4]), get(col_square[5]), get(col_square[6])
))

print(simple_cols)
```

### Map to the image

Now we can apply the simple colour set to the test image using `image_map()`. I've put the images side-by-side underneath using `image_append()`, also from `magick`.

```{r map}
# Map the simple colours to the image
duplo_mapped <- image_map(
  image = duplo,  # original image
  map = simple_cols  # colours to map on
)

# Display the original and simplified images side by side
image_animate(c(duplo, duplo_mapped), fps = 1)
```

Great. You can see where the original colours have been replaced by the 'closest' simple colours. Note in particular where the more reflective surfaces are mapped to white than the actual brick colour. This is okay: the brick may be blue, but  we've only defined _one_ shade of blue. If a particular shade is closer to white, then so be it.

## Quantify the colours

Now we can take this mapped image and quantify how much of the image belongs to each colour. Imagine we've broken the image into pixels and then counting how many belng to each of our six colours.

```{r count-colours}
# Function to count the colours (adapted from Jeroen Ooms)
count_colors <- function(image) {
  data <- image_data(image) %>%
    apply(2:3, paste, collapse = "") %>% 
    as.vector %>% table() %>%  as.data.frame() %>% 
    setNames(c("hex", "freq"))
  data$hex <- paste("#", data$hex, sep="")
  return(data)
}

# Dataframe of dominant colours 
duplo_col_freq <- duplo_mapped %>%
  count_colors() %>%
  left_join(
    as_tibble(cols_vec) %>%
      rownames_to_column() %>%
      rename(hex = value, name = rowname),
    by = "hex"
  ) %>% 
  arrange(desc(freq)) %>% 
  mutate(percent = 100*round((freq/sum(freq)), 3)) %>% 
  select(
    `Colour name` = name, Hexadecimal = hex, `Frequency of colour` = freq,
    `Percent of image` = percent
  )

duplo_mapped  # see mapped image again
kable(duplo_col_freq)  # quantify colour
```

So red makes up almost a quarter with white and black just behind. Many of the bricks are red and much of the shadow areas of yellow bricks were rendered as red. Black and white make up much of the shadowed and reflective surfaces elsewhere in the image.

# A few more examples

## Reef fish

[By Richard L Pyle from Wikimedia Commons, CC0 1.0.](https://commons.wikimedia.org/wiki/File:100%25_reef-fish_Endemism_at_90_m_off_Kure_Atoll.jpg)

```{r reef, echo=FALSE}
reef_path <- "https://upload.wikimedia.org/wikipedia/commons/0/05/100%25_reef-fish_Endemism_at_90_m_off_Kure_Atoll.jpg"

reef <- 
  image_read(reef_path) %>%
  image_scale(geometry = c("x600"))

reef_mapped <- image_map(
  image = reef,  # original image
  map = simple_cols  # colours to map on
)

reef_col_freq <- reef_mapped %>%
  count_colors() %>%
  left_join(
    as_tibble(cols_vec) %>%
      rownames_to_column() %>%
      rename(hex = value, name = rowname),
    by = "hex"
  ) %>% 
  arrange(desc(freq)) %>% 
  mutate(percent = 100*round((freq/sum(freq)), 3)) %>% 
  select(
    `Colour name` = name, Hexadecimal = hex, `Frequency of colour` = freq,
    `Percent of image` = percent
  )

image_animate(c(reef, reef_mapped), fps = 1)

kable(reef_col_freq)
```

## Neon lights in Hong Kong

[By Daniel Case from Wikimedia Commons, CC BY-SA 3.0](https://commons.wikimedia.org/wiki/File:Neon_lights,_Nathan_Road,_Hong_Kong.jpg)

```{r neon-lights, echo=FALSE}
neon_path <- "https://upload.wikimedia.org/wikipedia/commons/thumb/b/b0/Neon_lights%2C_Nathan_Road%2C_Hong_Kong.jpg/900px-Neon_lights%2C_Nathan_Road%2C_Hong_Kong.jpg"

neon <- 
  image_read(neon_path) %>%
  image_scale(geometry = c("x600"))

neon_mapped <- image_map(
  image = neon,  # original image
  map = simple_cols  # colours to map on
)

neon_col_freq <- neon_mapped %>%
  count_colors() %>%
  left_join(
    as_tibble(cols_vec) %>%
      rownames_to_column() %>%
      rename(hex = value, name = rowname),
    by = "hex"
  ) %>% 
  arrange(desc(freq)) %>% 
  mutate(percent = 100*round((freq/sum(freq)), 3)) %>% 
  select(
    `Colour name` = name, Hexadecimal = hex, `Frequency of colour` = freq,
    `Percent of image` = percent
  )

image_animate(c(neon, neon_mapped), fps = 1)

kable(neon_col_freq)
```

## Ladybird

[By Elena Andreeva from Wikimedia Commons, CC0 1.0.](https://commons.wikimedia.org/wiki/File:Erysimum_Cheiranthoides_(215134987).jpeg)

```{r bfly, echo=FALSE}
bfly_path <- "https://upload.wikimedia.org/wikipedia/commons/d/d5/Erysimum_Cheiranthoides_%28215134987%29.jpeg"

bfly <- 
  image_read(bfly_path) %>%
  image_scale(geometry = c("x600"))

bfly_mapped <- image_map(
  image = bfly,  # original image
  map = simple_cols  # colours to map on
)

bfly_col_freq <- bfly_mapped %>%
  count_colors() %>%
  left_join(
    as_tibble(cols_vec) %>%
      rownames_to_column() %>%
      rename(hex = value, name = rowname),
    by = "hex"
  ) %>% 
  arrange(desc(freq)) %>% 
  mutate(percent = 100*round((freq/sum(freq)), 3)) %>% 
  select(
    `Colour name` = name, Hexadecimal = hex, `Frequency of colour` = freq,
    `Percent of image` = percent
  )

image_animate(c(bfly, bfly_mapped), fps = 1)

kable(bfly_col_freq)
```

# Session info

```{r session-info}
devtools::session_info()
```

[^1]: Just as well, because I'm colourblind.
[^2]: There are [five named versions of olive drab](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf) in R.
[^3]: [Photo by Arto Alanenpää](https://commons.wikimedia.org/wiki/Lego#/media/File:Lego_dublo_arto_alanenpaa_2.JPG), CC0-BY-4.0 from Wikimedia Creative Commons.
[^4]: Artefacts introduced during compression of PNGs and JPGs might mean that your set of six colours ends up being _more_ than six. It's preferable to generate our colour set within R, inside `image_graph()`, so that we have only our six defined colours.