---
title: The Mountain Goats with geniusr
author: Matt Dray
date: '2018-04-30'
slug: the-mountain-goats-with-geniusr
draft: true
categories:
  - R
  - pop culture
  - text analysis
tags:
  - dplyr
  - tidytext
  - geniusr
  - lyrics
  - the mountain goats
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warnings = FALSE, message = FALSE, error = FALSE)
```

[Genius](https://genius.com/) is a lyrics site. You can access its data via API using [R](https://www.r-project.org/about.html) with the [`genuisr`](https://github.com/ewenme/geniusr) package by Ewen Henderson ([web](https://ewen.io/), [Github](https://github.com/ewenme), [Twitter](https://twitter.com/ewen_)).

```{r packages}
library(geniusr)  # access lyrics data (from dev version)
library(dplyr)  # data manipulation
library(tidytext)  # text wrangling
library(purrr)  # functional programming
library(beepr)  # for alerts
library(ggplot2)  # plotting
```

# Set up a genius app 

To set up:

1. create a [Genius API client](https://genius.com/api-clients/new)
2. click 'generate access token' under 'client access token' to generate an access token
3. after `install.packages("geniusr"); library(geniusr); geniusr::genius_token()` you'll be promopted to enter the access token (if you need to change this, you can use `geniusr::genius_token(force = TRUE)` to be re-prompted for a new token, which you can regenerate from the [Genius site](https://genius.com/api-clients) as in step 2)

Now you can use the functions in the 'geniusr' package.

# The Mountain Goats

Find the artist ID for the band.

```{r get_artist_id}
artist_id <- geniusr::search_artist("the mountain goats") %>% 
  dplyr::select(artist_id) %>% 
  dplyr::pull()  # column to vector
```

The ID is `print(artist_id)`.

Get the metadata for the band.

```{r get_artist_meta}
artist_meta <- geniusr::get_artist_meta(artist_id = artist_id)

print(artist_meta)
```

# Get songs

Use the artist ID to obtain all the songs on Genius. Out of interest, we can see the songs with the highest number of annotations -- notes left by users -- on the site.

```{r get_songs}
# get all songs for a given artist id
songs <- geniusr::get_artist_songs(artist_id = artist_id)

# a random sample of the returned 
songs %>% 
  dplyr::sample_n(10)
```

We can also access a greater list of data for each song, including the album name and release date.

```{r get_song_metadata}

songs_meta <- map_df(
  songs$song_id,
  get_song_meta
)

# get_song_meta(song_id = 1449474)
# # get the metadata for each song (album, etc)
# songs_meta <- geniusr::get_song_meta(songs, select(songs, song_id))
# 
# # join the meta data to the song data
# artist_songs %>%
#   left_join(
#     x = songs,
#     y = songs_meta,
#     by = song_id,
#     suffix = c("_song", "_meta")
#   )
```

# Get lyrics

Scrape the lyrics for each of the songs URLs.

There are some songs that have no lyrics, so let's remove these first.

```{r remove_instrumentals}
songs <- songs %>% 
  dplyr::filter(
    !song_lyrics_url %in% c(
      "https://genius.com/The-mountain-goats--lyrics",
      "https://genius.com/The-mountain-goats-for-the-west-coast-dark-ambient-bedroom-warriors-lyrics",
      "https://genius.com/The-mountain-goats-grave-dust-lyrics",
      "https://genius.com/The-mountain-goats-scaling-the-well-lyrics",
      "https://genius.com/The-mountain-goats-vanishing-act-lyrics"
    )
  )
```

Now we'll iterate over all of the song lyrics URLs to retrieve the lyrics associated with each.

This gives a dataframe with a row per line of lyrics, with columns for song name, lyrics URL and artist name.

```{r purrr_lyrics}
lyrics <- map_df(
  songs$song_lyrics_url,
  scrape_lyrics_url
)

lyrics <- lyrics %>%
  group_by(song_name) %>% 
  mutate(line_number = row_number()) %>% 
  ungroup(); beepr::beep(5)

# check out a sample of lines
lyrics %>% 
  dplyr::sample_n(10) %>% 
  select(line, song_name)
```

# Text analysis

## Extract words

```{r word_per_row}
lyrics_tidy <- lyrics %>%
  unnest_tokens(word, line) %>% 
  anti_join(stop_words)

sample_n(lyrics_tidy, 10) %>% 
  select(word, song_name)
```

## Word frequency

```{r count_words}
lyrics_tidy %>%
  count(word, sort = TRUE) 
```

And as a plot.

```{r plot_word_freq}
lyrics_tidy %>%
  count(word, sort = TRUE) %>%
  filter(n > 100) %>%
  mutate(word = reorder(word, n)) %>%
  ggplot(aes(word, n)) +
  geom_col() +
  xlab(NULL) +
  coord_flip() +
  theme_minimal()
```

## Sentiment

To what extent to songs and albums differ in sentiment?

Join a lexicon of sentiment-labelled words to the lyrics.

### By song

Assign a sentiment score for each song as a whole and look at the most positive nad most negative; grouping.

### By album

## Topic modelling

Can we group songs by similar lyrical content?
