---
title: The NBA Northwest Division with gmapsdistance
author: Matt Dray
date: '2018-12-18'
slug: nba-travel
categories:
  - R
tags:
  - NBA
  - gmapsdistance
  - dplyr
  - rvest
  - stringr
draft: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
library(gmapsdistance)  # because the chunk it's in is eval=FALSE
# Remember to do set.api.key("YOUR-API-KEY") before serving the site
```

Matt Dray ([\@mattdray](https://www.twitter.com/mattdray))

![Jazz's Stockton to Malone for the dunk ([via Giphy](https://giphy.com/gifs/vintage-basketball-GOIl4CelYNleM))](https://media.giphy.com/media/GOIl4CelYNleM/giphy.gif)

# TL;DR

You can query the Google Maps API in R with the `gmapsdistance` package. I used it to find travel times between NBA teams and learnt that travelling half the width of a continent requires a lot of petrol.  

# The NBA

The USA is quite big.

So it makes sense that the USA's National Basketball Association (NBA) organised their teams geographically: two 'conferences' -- Western and Eastern -- with three 'divisions' of five teams each. Right?

Yes, for the Central division of the Eastern Conference. It contains the Chicago Bulls, Cleveland Cavaliers, Detroit Pistons, Indiana Pacers and Milwaukee Bucks -- cities clustered near Lakes Michigan and Erie.

But also no, not for the Northwest Division of the Western Conference. This stretches from Portland Trail Blazers in the Pacific Northwest to Oklahoma City Thunder, which is actually in the country's south-central region. Far from clustering, Utah Jazz and Colorado's Denver Nuggets sit on a line between these two. The fifth team? That's Minnesota Timberwolves, which by rights should be part of the Central Division

What does this mean for the travelling basketball fans?

# Travel time with `gmapsdistance`

The R package [`gmapsdistance`](https://github.com/rodazuero/gmapsdistance) wraps the [Google Maps](https://www.google.co.uk/maps) [API](https://medium.freecodecamp.org/what-is-an-api-in-english-please-b880a3214a82). You can use it to query origin and destination points for travel distance (metres) and duration (seconds) for several modes of travel: on foot, by car, by public transport, or by bicycle (for parts of North America only).

We need an API key for Google Maps. This helps prevent misuse of their service and tracks your access. Your first $300 is free.

The process is outlined in detail on [the Google developer pages](https://developers.google.com/maps/documentation/distance-matrix/get-api-key#key). [Go to the website](https://cloud.google.com/maps-platform/) and follow the instructions.

Keep this key safe for now; we'll read it in later.

# Get data

We can use [Hadley Wickham's `rvest` package](https://github.com/hadley/rvest) to scrape the [Wikipedia page for the NBA)](https://en.wikipedia.org/wiki/National_Basketball_Association), which contains a table of teams and locations. 

We'll start by setting up our workspace.

```{r load-packages, message=FALSE, warning=FALSE}
library(dplyr)  # data manipulation
library(rvest)  # web scraping
library(stringr)  # string manipulation
```

Then we'll create objects that define the HTML for the web page and the table we want. The latter is defined using a CSS selector or, as in our case, an xpath. Think of these as 'addresses' for HTML content. You can use the 'inspect' function in your browser or [the selectorgadget browser extension](https://selectorgadget.com/) to get this information. 

```{r set-scrape-variables}
# Read the HTML from a specified web URL
html <- read_html("https://en.wikipedia.org/wiki/National_Basketball_Association")

# Use the Scrapy tool or 'inspect' in your browser to get the table xpath
table_xpath <- '//*[@id="mw-content-text"]/div/table[4]'
```

Now we can collect the table and manipulate it.

We could use the latitude and longitude from this table as our origin and destination points, but the arena name, city and state will be enough. We can create a column in our data in the form `arena+city+state`, which is the format required by Google Maps.

```{r get-nba-table}
nba_table <-
  html_nodes(x = html, xpath = table_xpath) %>%  # scrape the HTML
  html_table(fill = TRUE, header = NA) %>%  # parse as table
  as.data.frame() %>%  # convert from list
  filter(Division == "Northwest") %>%  # NW division only
  mutate(
    Search = paste(Arena, City.State),  # combine location details
    Search = str_replace_all(Search, ",", ""),  # replace comma with blank
    Search = str_replace_all(Search, " ", "+"),  # replace space with plus
    `Team code` = case_when(
      Team == "Denver Nuggets" ~ "DEN",
      Team == "Minnesota Timberwolves" ~ "MIN",
      Team == "Oklahoma City Thunder" ~ "OKC",
      Team == "Portland Trail Blazers" ~ "POR",
      Team == "Utah Jazz" ~ "UTA"
    )
  ) %>% 
  select(Team, `Team code`, Arena, City = City.State, Search)

nba_table 
```

So now we have a simple table of teams, locations and a search string that we'll pass to `gmapsdistance()`.

# Calculate distances

The basic arguments to the `gmpasdistance()` function are the origin and destination (can be an address, postcode or latlong coordinates), the mode of travel (car, public transit, walking) and the return format (shape) of the data (each origin-destination pair per row, or a matrix). We want all possible combinations

We can make our key available in the package environment by using the `set.api.key()` function from the `gmapsdistance` package.

```{r set-api-key, eval=FALSE}
library(gmapsdistance)  # install with install.packages() first

# Set the API key
set.api.key("YOUR-API-KEY")
```

Now we can query Google Maps with the API key. I've explained the function arguments in the code comments.

```{r use-gmapsdistance, cache=TRUE}
nba_travel <- 
  gmapsdistance(
    key = get.api.key(),  # retrieve the API key we set in the step above
    origin = nba_table$Search,  # vector of locations from our dataframe
    destination = nba_table$Search,  # as above
    combinations = "all",  # all possible pairs of locations
    mode = "driving",  # rather than walking, public transport or cycling
    shape = "long"  # one row per location-pair
  )
```

# Output

What does the output look like?

```{r nw_nba_list}
glimpse(nba_travel)
```

It's a list object with three elements: `Time`, `Distance` and `Status`. Each list element is a dataframe with three columns:

* `or` -- the origin point
* `de` -- the destination point
* one of `Time` in seconds, `Distance` in metres or `Status` (whether the request was successful)

All our requests returned 'OK' for `Status` so there were no problems wth fetching the data.

# Results: travel distance

## 

The data is returned in a tidy data frame with one row per origin-destination. We can clean this up by matching three-letter team codes[^1] to the data, arranging by distance within origin location and removing rows that provide the distance from each team to themselves (i.e. zero kilometres).

```{r distance-long}
# Create a mini-lookup of search strings to three-letter team codes
nba_lookup <- select(nba_table, `Team code`, Search)

# Make table easier to read by joining team codes
nba_distance <- nba_travel$Distance %>%
  left_join(nba_lookup, by = c("or" = "Search")) %>% 
  left_join(
    nba_lookup, by = c("de" = "Search"), suffix = c(" (or)", " (de)")
  ) %>% 
  mutate(Distance = round(Distance/1000, 1)) %>% # m to km
  select(`Team code (or)`, `Team code (de)`, Distance)

# View table
nba_distance %>%
  arrange(`Team code (or)`, desc(Distance)) %>% 
  filter(Distance != 0)
```

## Total distance per team

We can use this to summarise the total distance each team must travel to each of the others.

```{r}
nba_distance %>% 
  group_by(`Team code (or)`) %>% 
  summarise(`Total distance` = sum(Distance)) %>% 
  arrange(desc(`Total distance`))
```

## Travel distance lookup

Another way to view this is to 'spread' the data from long to wide format to create a look-up between origin teams (rows) and destination teams (columns).

```{r distance-lookup, message=FALSE}
# So we can spread out data into 'wide format'
library(tidyr)

# Create easy look-up table for travel distances between teams
nba_distance %>% 
  spread(key = `Team code (de)`, value = Distance) %>% 
  rename(Team = `Team code (or)`)
```

# Results: travel time

We can do a similar thing for time and print it in hours.

```{r time, message=FALSE}
# Package for tidy time handling
library(lubridate)

# Create easy look-up table for travel time between teams
nba_travel$Time %>%
  left_join(nba_lookup, by = c("or" = "Search")) %>% 
  left_join(
    nba_lookup, by = c("de" = "Search"), suffix = c(" (or)", " (de)")
  ) %>% 
  mutate(Time = round(Time/60/60, 1)) %>% 
  select(`Team code (or)`, `Team code (de)`, Time) %>% 
  spread(key = `Team code (de)`, value = Time)
```

[^1]: Codes: Denver Nuggets (DEN), Minnesota Timberwolves (MIN), Oklahoma City Thunder (OKC), Portland Trail Blazers (POR), Utah Jazz (UTA)