---
title: The NBA Northwest Division with gmapsdistance
author: Matt Dray
date: '2018-12-18'
slug: nba-travel
categories:
  - R
tags:
  - NBA
  - gmapsdistance
  - dplyr
  - rvest
  - stringr
draft: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval=FALSE)
library(gmapsdistance)  # because the chunk it's in is eval=FALSE
api_key <- read.table("~/Desktop/api_key.txt")[1, 1]  # stored locally
set.api.key(api_key)  # set key
```

Matt Dray ([\@mattdray](https://www.twitter.com/mattdray))

![Jazz's Stockton to Malone for the dunk ([via Giphy](https://giphy.com/gifs/vintage-basketball-GOIl4CelYNleM))](https://media.giphy.com/media/GOIl4CelYNleM/giphy.gif)

# The NBA

The USA is quite big.

So it makes sense that the USA's National Basketball Association (NBA) organised their teams geographically: two 'conferences' -- Western and Eastern -- with three 'divisions' of five teams each, right?

Yes, for the Central division of the Eastern Conference. It contains the Chicago Bulls, Cleveland Cavaliers, Detroit Pistons, Indiana Pacers and Milwaukee Bucks -- cities clustered near Lakes Michigan and Erie.

But also no, not for the Northwest Division of the Western Conference. This stretches from Portland Trail Blazers in the Pacific Northwest to Oklahoma City Thunder, which is actually in the country's south-central region. Far from clustering, Utah Jazz and Colorado's Denver Nuggets sit on a line between these two. The fifth team? That's Minnesota Timberwolves, which could be mistaken for part of the Central Division

![NBA team locations with states coloured by division (CC BY-SA 2.5, [Wikipedia](https://en.wikipedia.org/wiki/National_Basketball_Association#/media/File:Map_of_USA_and_Canada,_NBA,_zoom.svg))](/post/2018-12-18-googlemaps-in-the-nba-pacific-division_files/nba-map-wiki.png)

What does this mean for the travelling basketball fans?

# Google Maps and `gmapsdistance`

If you didn't know, Google Maps is one of many services that calculates journey distances and durations. 

The R package [`gmapsdistance`](https://cran.r-project.org/web/packages/gmapsdistance/index.html) was written by Rodrigo Azuero, Demetrio Rodriguez and David Zarruk. It wraps the [Google Maps](https://www.google.co.uk/maps) [API](https://medium.freecodecamp.org/what-is-an-api-in-english-please-b880a3214a82). You can use it to query origin and destination points for travel distance and duration for several modes of travel.

We need an API key for Google Maps. This helps prevent misuse of the service and helps keeps track of your credit. The process is outlined in detail in [the Google developer documentation pages](https://developers.google.com/maps/documentation/distance-matrix/get-api-key#key). [Go to the Google Cloud maps platform pages](https://cloud.google.com/maps-platform/) and follow the instructions. You'll need a Google account to do this.

The key is kept in [your Google Cloud console](https://console.cloud.google.com/).

[INSERT AN IMAGE FROM THE GOOGLE MAPS API SIGN-UP PROCESS]

# Get data

Let's use [Hadley Wickham's `rvest` package](https://github.com/hadley/rvest) to scrape the [Wikipedia page for the NBA)](https://en.wikipedia.org/wiki/National_Basketball_Association), which contains a table of teams and locations.

Start by loading `rvest` and some other packages for data manipulation.

```{r load-packages, message=FALSE, warning=FALSE}
library(dplyr)  # data manipulation and %>%
library(tidyr)  # tidying dataframes
library(stringr)  # string manipulation
library(rvest)  # web scraping
library(knitr)  # print tables nicely
```

Then we'll create objects that define the HTML for the web page and the table we want. The table can be defined by a CSS selector or an xpath; think of these as 'addresses' for HTML content. You can use the 'inspect' function in your browser or [the selectorgadget browser extension](https://selectorgadget.com/) to get this information.

```{r set-scrape-variables}
# Read the HTML from a specified web URL
html <- read_html("https://en.wikipedia.org/wiki/National_Basketball_Association")

# Use selectorgadget or your browser's 'inspect' mode to get xpath
table_xpath <- '//*[@id="mw-content-text"]/div/table[4]'
```

[INSERT BROSWER INSPECT WITH XPATH HIGHLIGHTED]

Now we can collect the table and get the parts we want: team names and locations. We can get the latter as both latitude-longitude coordinates and as text. Text can be parsed by Google Maps if words are separated by the plus symbol: `arena+city+state`.

Steps to do this are annotated in the code below.

```{r get-nba-table}
nba_table <-
  # Read HTML, isolate table, focus on NW
  html_nodes(x = html, xpath = table_xpath) %>%  # scrape the HTML
  html_table(fill = TRUE, header = NA) %>%  # parse as table
  as.data.frame() %>%  # convert from list
  filter(Division == "Northwest") %>%  # NW division only
  # Isolate the latitude and longitude
  separate(Coordinates, c("Coords1", "Coords2", "Coords3"), " / ") %>% 
  separate(Coords3, c("Latitude", "Longitude"), sep = "; ") %>% 
  separate(Longitude, c("Longitude", "X"), sep = " \\(") %>% 
  mutate(
    # Make coordinates numeric
    Latitude = as.numeric(Latitude),  # make numeric
    Longitude = stri_escape_unicode(Longitude),  # there's a weird character
    Longitude = str_replace(Longitude, "\\\\ufeff", ""),  # remove it
    Longitude = as.numeric(Longitude),  # make numeric
    # Create text search string
    Search = paste(Arena, City.State),  # combine location details
    Search = str_replace_all(Search, ",", ""),  # replace comma with blank
    Search = str_replace_all(Search, " ", "+"),  # replace space with plus
    # Add three-letter team codes to help simplify outputs
    `Team code` = case_when(
      Team == "Denver Nuggets" ~ "DEN",
      Team == "Minnesota Timberwolves" ~ "MIN",
      Team == "Oklahoma City Thunder" ~ "OKC",
      Team == "Portland Trail Blazers" ~ "POR",
      Team == "Utah Jazz" ~ "UTA"
    )
  ) %>% 
  select(  # Retain columns of interest only
    Team, `Team code`,
    Arena, City = City.State,
    Latitude, Longitude, Search
  )

kable(nba_table)  # print
```

# Map locations

We can map the latitude and longitude to prove we've isolated them correctly.

```{r map}
library(leaflet)  # for interactive maps

leaflet(nba_table) %>%
  addProviderTiles(providers$OpenStreetMap) %>%
  addAwesomeMarkers(
    lng = ~Longitude, lat = ~Latitude,
    popup = ~paste0(
      "<b>", nba_table$Team, "</b>",
      "<br>", nba_table$Arena,
      "<br>", nba_table$City
    ),
    icon = awesomeIcons(
      library = "ion", icon = "ion-ios-basketball", # add icon
      markerColor = case_when(
        nba_table$`Team code` == "DEN" ~ "darkblue",
        nba_table$`Team code` == "MIN" ~ "darkblue",
        nba_table$`Team code` == "OKC" ~ "lightblue",
        nba_table$`Team code` == "POR" ~ "black",
        nba_table$`Team code` == "UTA" ~ "darkblue"
      ),
      iconColor = case_when(
        nba_table$`Team code` == "DEN" ~ "yellow",
        nba_table$`Team code` == "MIN" ~ "lightblue",
        nba_table$`Team code` == "OKC" ~ "orange",
        nba_table$`Team code` == "POR" ~ "red",
        nba_table$`Team code` == "UTA" ~ "green"
      )
    )
  ) %>%
  addMeasure()
```

# Calculate distances

The basic arguments to the `gmpasdistance()` function are the origin and destination (can be an address, postcode or latlong coordinates), the mode of travel (car, public transit, walking) and the return format (shape) of the data (each origin-destination pair per row, or a matrix). We want all possible combinations

We can make our key available in the package environment by using the `set.api.key()` function from the `gmapsdistance` package.

```{r set-api-key, eval=FALSE}
library(gmapsdistance)
set.api.key("YOUR-API-KEY")
```

Now we can query Google Maps with the API key. I've explained the function arguments in the code comments.

```{r use-gmapsdistance, cache=TRUE}
nba_travel <- 
  gmapsdistance(
    key = get.api.key(),  # retrieve the API key we set in the step above
    origin = nba_table$Search,  # vector of locations from our dataframe
    destination = nba_table$Search,  # as above
    combinations = "all",  # all possible pairs of locations
    mode = "driving",  # rather than walking, public transport or cycling
    shape = "long"  # one row per location-pair
  )
```

# Output

What does the output look like?

```{r nw_nba_list}
glimpse(nba_travel)
```

It's a list object with three elements: `Time`, `Distance` and `Status`. Each list element is a dataframe with 25 observations (each team paired with every other) and 3 columns:

* `or` -- the origin point
* `de` -- the destination point
* one of `Time` in seconds, `Distance` in metres or `Status` (whether the request was successful)

All our requests returned 'OK' for `Status` so there were no problems wth fetching the data.

# Results: travel distance

The data is returned in a tidy data frame with one row per origin-destination. We can clean this up by matching three-letter team codes to the data, arranging by distance within origin location and removing rows that provide the distance from each team to themselves (i.e. zero kilometres).

```{r distance-long, warning=FALSE}
# Create a mini-lookup of search strings to three-letter team codes
nba_lookup <- select(nba_table, `Team code`, Search)

# Make the output table easier to read by joining team codes
nba_distance <- nba_travel$Distance %>%
  left_join(nba_lookup, by = c("or" = "Search")) %>% 
  left_join(
    nba_lookup, by = c("de" = "Search"), suffix = c(" (or)", " (de)")
  ) %>% 
  mutate(Distance = round(Distance/1000, 1)) %>% # m to km
  select(`Team code (or)`, `Team code (de)`, Distance)

# View table
nba_distance %>%
  arrange(`Team code (or)`, desc(Distance)) %>% 
  filter(Distance != 0) %>% 
  kable()
```

# Travel distance summaries

It's trivial to summarise the furthest travel distance within the Northwest for each team.

```{r furthest-game}
nba_distance %>% 
  arrange(`Team code (or)`, desc(Distance)) %>%
  group_by(`Team code (or)`) %>% 
  slice(1) %>% 
  ungroup() %>% 
  arrange(desc(Distance)) %>% 
  kable()
```

Similarly, we can get the total distance each team must travel to each of the others in the Northwest Division.

```{r total-distance}
nba_distance %>% 
  group_by(`Team code (or)`) %>% 
  summarise(`Total distance` = sum(Distance)) %>% 
  arrange(desc(`Total distance`)) %>% 
  kable()
```

## Travel distance lookup

Another way to view this is to 'spread' the data from long to wide format to create a look-up between origin teams (rows) and destination teams (columns).

```{r distance-lookup, message=FALSE}
# Create easy look-up table for travel distances between teams
nba_distance %>% 
  spread(key = `Team code (de)`, value = Distance) %>% 
  rename(Team = `Team code (or)`) %>% 
  kable()
```

# Results: travel time

You can do all of the above for travel time as well. I'm going to provide the simple lookup table for thr sake of brevity.

```{r time, warning=FALSE, message=FALSE}
library(lubridate)  # tidy time handling

# Create easy look-up table for travel time between teams
nba_travel$Time %>%
  left_join(nba_lookup, by = c("or" = "Search")) %>% 
  left_join(
    nba_lookup, by = c("de" = "Search"), suffix = c(" (or)", " (de)")
  ) %>% 
  mutate(Time = round(Time/60/60, 1)) %>% 
  select(`Team code (or)`, `Team code (de)`, Time) %>% 
  spread(key = `Team code (de)`, value = Time) %>% 
  kable()
```