---
title: R session info info
author: Matt Dray
date: '2018-10-09'
slug: sessioninfo
categories:
  - R
tags:
  - devtools
  - sessioninfo
  - xfun
  - utils
draft: true
---

# What's session info?

'Session info' is just a print-out of some information about your R version, the locale you're running it from and the packages you've attached.

You might get asked for this if someone is diagnosing your R problem. You may also see it at the end of blog posts or reports.

The typical method is to run `sessionInfo()`, but it gives a *lot* of information by default. Other functions have been created to make the output more friendly.

# Why?

I've been using `sessionInfo` in my posts, but read about [Yihui's `session_info()` function in his `xfun` package](https://yihui.name/xfun/) and saw [Mat](https://twitter.com/mammykins_) at [machinegurning.com](https://www.machinegurning.com/rstats/bayes_thinking/) using `session_info()` from `devtools`.

I just want to see the outputs side-by-side to sate my own curiosity, to be honest.

# Load packages

Let's grab some packages from multiple sources to see how they're represented in the output of each session info function.

We'll get some packages from:

* [CRAN](https://cran.r-project.org/)
* [Bioconductor](https://www.bioconductor.org/install/)
* [GitHub](http://kbroman.org/pkg_primer/pages/github.html)

```{r eval=FALSE}
library("tidyxl")

source("https://bioconductor.org/biocLite.R")
biocLite()
biocLite(c("GenomicFeatures", "AnnotationDbi"))

devtools::install_github("tidyverse/dplyr")
```

# The functions

## `utils`

From the [documentation for the function](https://www.rdocumentation.org/packages/utils/versions/3.5.1/topics/sessionInfo):

> Print version information about R, the OS and attached or loaded packages.

```{r}
utils::sessionInfo()
```

## `sessioninfo` package

From [the CRAN page for the package](https://CRAN.R-project.org/package=sessioninfo):

> It is similar to `utils::sessionInfo()`, but includes more information about packages, and where they were installed from.

```{r}
sessioninfo::session_info()
```

## `devtools` package

From the [documentation for the function](https://www.rdocumentation.org/packages/devtools/versions/1.13.6/topics/session_info):

> This is `sessionInfo()` re-written from scratch to both exclude data that's rarely useful (e.g., the full collate string or base packages loaded) and include stuff you'd like to know (e.g., where a package was installed from).

```{r}
devtools::session_info()
```

## `xfun` package

From [Yihui Xie's Introduction to `xfun` page](https://yihui.name/xfun/):

>...it is rarely useful to print out the names of base R packages, or information about the matrix products / BLAS / LAPACK. Often times I want additional information in the session information, such as the Pandoc version when rmarkdown is used.

```{r}
xfun::session_info()
```