---
title: The NBA Northwest Division with gmapsdistance
author: Matt Dray
date: '2018-12-18'
slug: nba-travel
categories:
  - R
tags:
  - NBA
  - gmapsdistance
  - dplyr
  - rvest
  - stringr
draft: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
library(gmapsdistance)  # because the chunk it's in is eval=FALSE
# Remember to do set.api.key("YOUR-API-KEY") before serving the site
```

Matt Dray ([\@mattdray](https://www.twitter.com/mattdray))

# TL;DR

You can query the Google Maps API in R with the `gmapsdistance` package. I used it to find travel times between NBA teams and learnt that travelling half the width of a continent requires a lot of petrol.  

# The NBA

The USA is quite big.

So it makes sense that the USA's National Basketball Association (NBA) organised their teams geographically: two 'conferences' -- Western and Eastern -- with three 'divisions' of five teams each. Right?

Yes, for the Central division of the Eastern Conference. It contains the Chicago Bulls, Cleveland Cavaliers, Detroit Pistons, Indiana Pacers and Milwaukee Bucks -- cities clustered by the Great Lakes.

But also no, not for the Northwest Division of the Western Conference. This stretches from Portland Trail Blazers in the Pacific Northwest to Oklahoma City Thunder, which is actually in the country's south-central region. Far from clustering, Utah Jazz and Colorado's Denver Nuggets sit on a line between these two. The fifth team? That's Minnesota Timberwolves, which by rights should be part of the Central Division

What does this mean for the travelling basketball fans?

# Travel time with `gmapsdistance`

The R package [`gmapsdistance`](https://github.com/rodazuero/gmapsdistance) wraps the [Google Maps](https://www.google.co.uk/maps) [API](https://medium.freecodecamp.org/what-is-an-api-in-english-please-b880a3214a82). You can use it to query origin and destination points for travel distance (metres) and duration (seconds) for several modes of travel: on foot, by car, by public transport, or by bicycle (for parts of North America only).

We need an API key for Google Maps. This helps prevent misuse of their service and tracks your access. Your first $300 is free.

The process is outlined in detail on [the Google developer pages](https://developers.google.com/maps/documentation/distance-matrix/get-api-key#key). [Go to the website](https://cloud.google.com/maps-platform/) and follow the instructions.

Keep this key safe for now; we'll read it in later.

# Get data

We can use [Hadley Wickham's `rvest` package](https://github.com/hadley/rvest) to scrape the [Wikipedia page for the NBA)](https://en.wikipedia.org/wiki/National_Basketball_Association), which contains a table of teams and locations. 

We'll start by setting up our workspace.

```{r load-packages, message=FALSE, warning=FALSE}
library(dplyr)  # data manipulation
library(rvest)  # web scraping
library(stringr)  # string manipulation
```

Then we'll create objects that define the HTML for the web page and the table we want. The latter is defined using a CSS selector or, as in our case, an xpath. Think of these as 'addresses' for HTML content. You can use the 'inspect' function in your browser or [the selectorgadget browser extension](https://selectorgadget.com/) to get this information. 

```{r set-scrape-variables}
# Read the HTML from a specified web URL
html <- read_html("https://en.wikipedia.org/wiki/National_Basketball_Association")

# Use the Scrapy tool or 'inspect' in your browser to get the table xpath
table_xpath <- '//*[@id="mw-content-text"]/div/table[4]'
```

Now we can collect the table and manipulate it.

We could use the latitude and longitude from this table as our origin and destination points, but the arena name, city and state will be enough. We can create a column in our data in the form `arena+city+state`, which is the format required by Google Maps.

```{r get-nba-table}
nba_table <-
  html_nodes(x = html, xpath = table_xpath) %>%  # scrape the HTML
  html_table(fill = TRUE, header = NA) %>%  # parse as table
  as.data.frame() %>%  # convert from list
  filter(Division == "Northwest") %>%  # NW division only
  mutate(
    Search = paste(Arena, City.State),  # combine location details
    Search = str_replace_all(Search, ",", ""),  # replace comma with blank
    Search = str_replace_all(Search, " ", "+")  # replace space with plus
  ) %>% 
  select(Team, Arena, State = City.State, Search)  # columns of interest
```

So now we have a simple table of teams, locations and a search string that we'll pass to `gmapsdistance()`.

# Calculate distances

The basic arguments to the `gmpasdistance()` function are the origin and destination (can be an address, postcode or latlong coordinates), the mode of travel (car, public transit, walking) and the return format (shape) of the data (each origin-destination pair per row, or a matrix). We want all possible combinations

We can make our key available in the package environment by using the `set.api.key()` function from the `gmapsdistance` package.

```{r set-api-key, eval=FALSE}
library(gmapsdistance)  # install with install.packages() first

# Set the API key
set.api.key("YOUR-API-KEY")
```

Now we can query Google Maps with the API key.

```{r use-gmapsdistance, cache=TRUE}
nba_travel <- 
  gmapsdistance(
    key = get.api.key(),  # retrieve API key that we set above
    origin = nba_table$Search,  # start point of journey
    destination = nba_table$Search,  # end point of journey
    mode = "driving",  # driving time
    combinations = "all",  # all possible pairs of locations
    shape = "long"  # format of output data (origin and destination as cols)
  )
```

# Output

What does the output look like?

```{r nw_nba_list}
glimpse(nba_travel)
```

It's a list object with three elements:

1. Time in seconds
2. Distance in metres
3. Status of the call -- was it successful?

Each list element has three columns:

* `or` -- the origin point
* `de` -- the destination point
* one of `Time`, `Distance` and `Status`

Let's start with `Status`. Were all the requests actioned?

```{r status}
sch_distances$Status  # isolate status element of returned list
```

We want the status `OK`, which indicates that there were no problems and the distances were collected with no errors. The `PLACE_NOT_FOUND` error is returned in the `Status` column when Google Maps can't locate your origin or destination.

So what were the distances between the locations in metres?

```{r distance}
sch_distances$Distance  # isolate distance (metres) element of returned list
```

And how much time does this translate to, in seconds, when driving between the locations?

```{r time}
sch_distances$Time  # isolate time (seconds) element of returned list
```

# As a single dataframe

That's great, but still not super-friendly to use, especially over long distances.

# Findings

While we're at it, we can look arbitrarily at the longest distances.

```{r arrange}
distance_info %>%
  dplyr::mutate(  # create new columns
    Kilometres = round(Distance/1000, 1),  # calculate km from m
    Miles = round(Kilometres * 0.621371, 1)  # convert to miles
  ) %>% 
  dplyr::select(  # select columns to rename and retain
    Origin = establishmentname_or,
    Destination = establishmentname_de,
    `Kilometres`,
    `Miles`
    ) %>% 
  dplyr::arrange(desc(Kilometres)) # arrange by longest distance first
```
