---
title: The Mountain Goats with {trelliscopejs}
author: Matt Dray
date: '2019-06-19'
slug: goat-scope
categories:
  - R
  - dataviz
tags:
  - trelliscopejs
  - trelliscope
  - small multiples
draft: yes
---


# Small multiples

The [{trelliscopejs}]() package by [Ryan Hafen]() harnesses the power of his [trelliscope.js]() JavaScript library for R.

What does it do? It provides an interactive interface for organising and exploring visualisations of small mulitples.

What's a 'small multiple'? Rather than overplotting data for multiple levels of some variable, you can facet by them and display the outputs side by side for easy comparison.

Ryan has written [guidance]() and provided examples in posts using data on [x](), [Pokemon]() and [Instagram]().[^education] His package is relatively simple to use and does a lot of legwork for you to provide a nice interface to your data. 

# Goat discography

In a previous post I used the {spotifyr}, {genius} and {markovifyR} packages to generate new lyrics for the band The Mountain Goats.

The data from Spotify is interesting. It has musical information like key and tempo, but also audio features like 'danceability' and 'acousticness' scaled from 0 to 1. We also got links to album art in that dataset.

I'm going to use these data in this post to provide a Trelliscope example.  

# Ready the data

We'll load {trelliscopejs} and some [tidyverse packages]() to help us out.

```{r packages, warnings=FALSE, message=FALSE}
library(trelliscopejs)
library(dplyr)  # manipulate data
library(readr)  # read data
library(tidyr) # get data into tidy form
```

## Get data and simplify

You can follow the instructions in the previous post to get the data. I'm going to load those data from an RDS file on my machine.

```{r read-data}
raw_goat <- read_rds("~/Desktop/goat_discography.RDS")
dim(raw_goat)
```

It's a pretty large data set, so let's simplify.

```{r}
small_goat <- raw_goat %>% 
  unnest(available_markets) %>%  # unnest character vector
  filter(available_markets == "US") %>%  # releases from one country only
  select(
    track_name, album_name, track_n, album_release_year,  # track detail
    danceability, energy, speechiness, acousticness,  # audio features
    instrumentalness, liveness, valence  # audio features
  ) %>%
  arrange(energy)  # order by 'energy' audio feature

glimpse(small_goat)
```

## Album artwork

Unnesting the `available_markets` character vector removed the `album_image` variable, which is a nested dataframe with URLs to different-sized album artwork. We can grab unique album image URLs and join them back to our dataset.

```{r image-unnest}
goat_pics <- raw_goat %>%
  unnest(album_images) %>%  # unnest dataframe fo URLs
  filter(width == 640) %>%  # just the largest images
  select(album_name, url) %>%  # simplify dataset
  distinct(album_name, .keep_all = TRUE)  # one unique entry per album

glimpse(goat_pics)
```

And now to join the album image URLs back to our simplified dataset.

```{r join}
small_goat_pics <- left_join(
  x = small_goat,
  y = goat_pics,
  by = "album_name"
)

glimpse(small_goat_pics)
```

So the album artwork URL has been added to the `url` column.

## Prep for Trelliscope

This gives us a nice tidy dataframe, but I'm going to make a couple more changes to prepare the data for Trelliscope. First, I need to use the `img_panel()` function to declare that the album images should be the thing being visualised. Then I can rename the variables to make them look nicer when displayed.

```{r panel}
prepared_goat <- small_goat_pics %>% 
  mutate(panel = img_panel(url)) %>%  # identify as viz for panel
  rename_all(tools::toTitleCase) %>%  # first letter capitalised
  rename(
    Track = Track_name,
    Album = Album_name,
    `Track #` = Track_n,
    Year = Album_release_year
  ) %>% 
  select(-Url)  # discard unneeded variable
```

# Generate Trelliscope

Now we're ready. The call to `trelliscope()` takes the dataset and then a bunch of other arguments like a title and subtitle and the default state for the number of rows and columns of panels and the default data to show on the panel under the visualisation.

```{r eval=FALSE}
trelliscope(
  prepared_goat,
  name = "The Mountain Goats discography",
  desc = "Explore the Mountain Goats backcatalogue and filter by Spotify metrics like danceability and acousticness",
  nrow = 2, ncol = 5,  # arrangement of panels
  state = list(labels = c("Track", "Album", "Track #", "Year", "Energy"))  # display on panels
)
```

I've embedded here the output, but [you can click here to open full screen](https://www.rostrum.blog/output/goat_scope/).

```{r include-url, echo=FALSE}
knitr::include_url("https://www.rostrum.blog/output/goat_scope/")
```

Explore the data by altering the defaults in the grid, labels, filter and sort buttons in the left-hand navigation panel. Cycle through the panels with the arrows in the upper right.

# Bonus content! 

## Energy

By default, the panels in this trelliscope are ordered by the 'energy' of a track; most energetic first. The top track for energy is one of my favourites: 'Choked Out' from 'Beat the Champ'. Here's a Spotify snippet.

<iframe src="https://open.spotify.com/embed/track/6VAOhkoZypZtGFgrnIGr48" width="300" height="380" frameborder="0" allowtransparency="true" allow="encrypted-media"></iframe>

## Host your trelliscope

Use the argument `path = "/tmp/goat-scope"` in the `trelliscope()` functino to save the files for your trelliscope to the temporary folder. You can then host the folder's contents somewhere. I put mine in the `static/` directory of this site's structure so it's hosted on its own page at https://www.rostrum.blog/output/goat_scope/.

[^education]: I also once explored the use of [Trelliscope for UK education data]() and have been meaning to write about it ever since.