---
title: Teach a human to swirl
author: Matt Dray
date: '2019-04-30'
slug: teach-a-human-to-swirl
categories:
  - R
  - teaching
tags:
  - swirl
  - swirlify
draft: yes
---

Matt Dray ([\@mattdray](https://www.twitter.com/mattdray))

# tl;dr

1. [{swirl}](https://swirlstats.com/) lets people learn R from within R.
1. [{swirlify}](http://swirlstats.com/swirlify/) is a package to help write Swirl courses
1. Make a Swirl course -- I've begun [Tidyswirl](https://github.com/matt-dray/tidyswirl) for teaching [tidyverse](https://www.tidyverse.org/) packages

# Swirl

The {swirl} package lets you learn R interactively from within R itself.

Swirl prints questions in the console and users have to answer correctly before they can move on. 

To begin, install and load the package then follow the instructions that Swirl gives you.

```{r, install-swirl, eval=FALSE}
install.packages("swirl")
library(swirl)
swirl()
```

Full instructions are on Swirl's [website]().

New courses can be downloaded from the [Swirl Course Network](), but can also be sourced from GitHub or a zip file, for example.

I like Swirl. It's active, self-paced and extensible. But this post isn't about using Swirl, it's about creating new Swirl courses.

# Teach others to teach

Swirl is a fish. Teaching people to fish is Swirlify. Teaching people to teach people to fish is this post.

Swirl suffers because its library of courses is relatively small. More people need to make materials and share them. This shouldn't be a problem for the thriving R community.

## Swirlify

Sean Kross and Nick Carchedi have created [the {swirlify} package](https://CRAN.R-project.org/package=swirlify) to make course creation easy.

[Full Swirlify guidance](http://swirlstats.com/swirlify/) is available. This post covers the basics to get you started.

## Simple workflow

Let's create a new course about insects where each lesson is about a different order. To start:

```{r, new-lesson, eval=FALSE}
new_lesson(course_name = "Course", lesson_name = "Lesson 1")
```

These folders and template files get created:

```
Home
└── Course
    └── Lesson_1
        ├── customTests.R
        ├── dependson.txt
        ├── initLesson.R
        └── lesson.yaml
```

Lesson content goes in `lesson.yaml`. This is the only file we care about right now. But we don't have to touch it. Begin content creation with:

```{r, swirlify, eval=FALSE}
swirlify()
```

The Shiny app GUI will open. On the left is where you create new questions. On the right you can see the lesson file itself.

To build a question:

1. Choose a question type from the dropdown menu
1. Fill in the text boxes ('Make Answer Test from Correct Answer' will help you by autofilling 'Answer Tests' from 'Correct Answer')
1. Click 'Add Question'
1. Click 'Save Lesson'
1. optional: click 'Demo Lesson' to try it out

Rinse and repeat until you've finished the lesson.

To create the next lesson, use the code above, but change the `lesson_name` argument to the name of the new lesson.

## Share it

There are several ways to share it.

## Minimal extra info

* Different question types
* 'Message' questions should be used to provide information -- they're statements rather than questions
* Answer tests
* YAML
* Auto-test your course

# Tidyswirl

I've started a Swirl course that might be useful. It's called Tidyswirl and the goal is to have a lesson that introduces each core package of [the tidyverse](https://www.tidyverse.org/).

Please feel free to contribute new courses or raise issues.

To install the course so you can use it in Swirl:

```{r, install-tidyswirl, eval=FALSE}
install_course_github("matt-dray", "tidyswirl")
```