---
title: Markov-chaining my PhD thesis II
author: Matt Dray
date: '2019-04-29'
slug: markov-chain-phd-2
categories:
  - R
  - text analysis
tags:
  - phd
  - markov chain
draft: yes
---

Matt Dray ([\@mattdray](https://www.twitter.com/mattdray))

I posted about using [a Markov chain](https://hackernoon.com/from-what-is-a-markov-model-to-here-is-how-markov-models-work-1ac5f4629b71) to generate sentences using [my PhD thesis](https://orca.cf.ac.uk/68365/) as input. I also posted about [the {markovifyR} package}(https://github.com/abresler/markovifyR) for [generating lyrics by The Mountain Goats](https://www.rostrum.blog/2019/04/25/gen-tmg-lyrics/).

So now I'm revisting the PhD post using {markovifyR} for my own amusement.

The PhD text is available from my {dray} package. I'll remove the blank lines and ignore the preamble and references section.

```{r}
library(dray)  # remotes::install_github("matt-dray/dray")
phd_text <- dray::phd  # get data
phd_text <- phd[nchar(phd) > 0]  # remove blank lines
phd_text <- phd[27:376]  # ignore preamble and references
```

Now to prepare the workspace, build the model and generate some sentences.

```{r}
# Prepare the workspace
# system("pip install markovify")
library(markovifyR)  # remotes::install_github("abresler/markovifyR")
library(furrr)  # install.packages("furrr")

# Build model
markov_model <- generate_markovify_model(
  input_text = phd,
  markov_state_size = 2L,
  max_overlap_total = 25,
  max_overlap_ratio = 0.7
)

# Generate lines
phd_speak <- markovify_text(
  markov_model = markov_model,
  maximum_sentence_length = NULL,
  output_column_name = 'phd_speak',
  count = 50,
  tries = 100,
  only_distinct = TRUE,
  return_message = TRUE
)
```

Some classics:

* Detritivorous macroinvertebrate species used to overcome this issue in future, where no material is used to hold them together.
This reduced decomposition rate in the University’s Open Access repository and for inter-library loan, and for inter-library loan, and for inter-library loans.
* Litter was produced under artificial conditions.
* Bonferroni-adjusted critical significance levels were compared by pairwise PERMANOVA.
* A GLMM was also affected by growth condition, to be made available to outside organisations.
* Canadian Journal of the laboratory in a randomly-ordered 3 × 3 grid.
* I am grateful for the choice test.
* This thesis is the total nitrogen content of the North American Benthological Society.
* In general, litter chemical composition and decomposition.
* This simplicity has allowed for broad underlying principles to be linked to litter chemical composition differently depending on whether it was exposed to a labeled 0.5 m steel rod.
* My parents have never wavered in their responses, suggesting that caution has to be available online in the CEF.
* Deciduous woodlands are dependent on the parent tree.
* After falling as litter, elevated-CO2 leaves are maintained after falling as litter.