---
title: Fix leaky {dplyr} pipes
author: Matt Dray
date: '2019-04-01'
slug: fix-leaky-dplyr-pipes
categories:
  - R
tags:
  - dplyr
  - tidylog
  - pipecleaner
  - magrittr
  - tidyverse
draft: yes
---

Matt Dray ([\@mattdray](https://www.twitter.com/mattdray))

## C'est un pipe

The `%>%` (pipe) operator lets you chain function calls. Stefan Milton Bache made it to 'decrease development time and to improve readability and maintainability of code'.[^vignette]

Voilà.

```{r pipeline-example}
# {dplyr} for data manipulation and to load pipe
suppressPackageStartupMessages(library(dplyr))

# Get mean sepal width for two iris species
iris_pipe <- iris %>%
  filter(Species %in% c("setosa", "versicolor")) %>% 
  group_by(Species) %>% 
  summarise(`Mean width` = mean(Sepal.Width)) %>% 
  mutate(`Mean width` = round(`Mean width`, 1))

# Print
iris_pipe
```

# Ce n'est pas debuggable?

[Some are critical of this approach](https://twitter.com/TedPetrou/status/1109519764613787648?ref_src=twsrc%5Etfw). Long pipes obscure what's happened to your data and make debugging hard.

This post summarises some solutions. This table summarises it even more.


| Package             | Shtick                                              | View     | Print    | Debug    |
| :------------------ | :-------------------------------------------------- | :------- | :------- | :------- |
| {tidylog}           | Prints human-readble results of each step           | &#10004; | &#10004; | &#10008; |
| {ViewPipeSteps}     | RStudio addin to reveal changes to data             | &#10004; | &#10004; | &#10008; |
| {pipecleaner}       | RStudio debug addin; helps 'burst' pipes            | &#10008; | &#10008; | &#10004; |
| {pipes}             | Operators for special assignment                    | &#10004; | &#10004; | &#10004; |
| {magrittr}          | `debug_pipe()` function                             | &#10004; | &#10004; | &#10004; |
| {tamper} (archived) | Stack trace replacement for pipe debugging          | &#10008; | &#10008; | &#10004; |
| [Bizarro pipe]      | Replace `%>%` with `->.;` and observe `.Last.level` | &#10008; | &#10008; | &#10008; 

# Ce n'est pas une probleme?

To be clear, I think most people create pipes interactively and check their outputs at each step.

You could also make sensibly-lengthed pipes that reflect a 'unit' of wrangling. Perhaps objects for the raw data as read-in, a reshaped version, one with cleaned columns, and another containing the model. You can then interrogate these easily.

Nevertheless.

# Fix the leaks

## {tidylog}

[The {tidylog} package](https://github.com/elbersb/tidylog) prints to the console the changes that have happened to your data after each {dplyr} step. You don't need to stop and print at every step.

```{r load-tidylog}
library(tidylog)  # must be loaded after dplyr
```

You can see {tidylog} masks the {dplyr} functions to print a human-readable summary sentence.

```{r tidylog-example}
iris_pipe <- iris %>%
  filter(Species %in% c("nope", "uh-uh")) %>% 
  group_by(Sepal.Width) %>% 
  summarise(`Mean width` = mean(Sepal.Width)) %>% 
  mutate(`Mean width` = round(`Mean width`, 1))
```

Did you spot the error? I filtered for species that don't exist in the dataset, so I `removed all rows (100%)` in the first step. 

I'll unload {tidylog} before continuing so it doesn't interfere with the other examples.

```{r tidylog-unload}
unloadNamespace("tidylog")
```

## {ViewPipeSteps}

[The {ViewPipeSteps} package](https://github.com/daranzolin/ViewPipeSteps) is an RStudio add-in. Select 'View/Print Pipe Chain Steps' form the add-ins menu.

Basically it runs `View()` or `print()` for each of the steps in your pipeline so you can see the the data. This is different to the summary sentence you get from {tidylog}, for example.

## {pipecleaner}

[The {pipecleaner} package](https://github.com/alistaire47/pipecleaner) is an RStudio addin. It has the best name. It's not on CRAN, so install with:

```{r install-pipecleaner}
devtools::install_github("alistaire47/pipecleaner")
```

You highlight your code and select 'Debug pipeline' from the Addins menu. The function does two things: first it 'bursts' your pipes (nice analogy) by breaking the pipeline into intermediate objects, one for each function that's contained within it. The interactive debug menu is then opened so you can execute your code step-by-step, see the results and perform other debug operations.

```{r burst-pipe}
# Intact
iris_pipe <- iris %>%
  filter(Species %in% c("nope", "uh-uh")) %>% 
  group_by(Sepal.Width) %>% 
  summarise(`Mean width` = mean(Sepal.Width)) %>% 
  mutate(`Mean width` = round(`Mean width`, 1))

# Burst
dot1 <- filter(iris, Species %in% c("setosa", "versicolor"))
dot2 <- group_by(dot1, Species)
dot3 <- summarise(dot2, `Mean width` = mean(Sepal.Width))
iris_pipe <- mutate(dot3, `Mean width` = round(`Mean width`,1))
```

You can choose 'Burst pipes' from the Addins menu without entering debug mode. 



## {magrittr}

[The {magrittr} package](https://magrittr.tidyverse.org/reference/debug_pipe.html) has its own `debug_pipe()` function that takes you into debug mode.

```{r magrittr-example}
library(magrittr)

test <- iris %>%
  filter(Species %in% c("setosa", "versicolor")) %>%
  group_by(Species) %>% 
  summarise(`Mean width` = mean(Sepal.Width)) %>% 
  mutate(`Mean width` = round(`Mean width`, 1)) %>% debug_pipe
```

<!-- TODO: below -->

## {pipes}

There's a few special operators in the {magrittr} package beyond `%>%`. For example, the `%T>%` (tee) operator outputs the original value, not the result of it.

X developed the {pipes} package to supply a bunch of other `%>%`-inspired operators.

devtools::install_github("moodymudskipper/pipes")

```{r pipes-example}
library(pipes)

iris_pipe <- iris %>%
  filter(Species %in% c("setosa", "versicolor")) %>% 
  group_by(Species) %>% 
  summarise(`Mean width` = mean(Sepal.Width)) %>%
  mutate(`Mean width` = round(`Mean width`, 1))

unloadNamespace("pipes")
```

## {tamper}

devtools::install_github("gaborcsardi/tamper")
https://github.com/gaborcsardi/tamper

```{r}
devtools::install_github("gaborcsardi/tamper")
library(tamper)
options(error = tamper::tamper)

iris_pipe <- iris %>%
  filter(Species %in% c("setosa", "versicolor")) %>% 
  group_by(Species) %>% 
  summarise(`Mean width` = mean(Sepal.Girth)) %>%
  mutate(`Mean width` = round(`Mean width`, 1))
```




## Bizarro pipe

We can hack our own 'operator' that acts like a pipe and can be run so that we can check what's happening at each step.

It's [a 'Bizarro pipe'](http://www.win-vector.com/blog/2017/01/using-the-bizarro-pipe-to-debug-magrittr-pipelines-in-r/): `->.;`. What it's saying is 'right-assign to a dot and then perform the next operation'.

Things you might be wondering:

* yes, you can use a `->` for assignment
* yes, you can assign to a `.`, but you'll need to explicitly supply it as the data argumnt to the next function call in your 'Bizarro pipeline'
* yes, you can use semi-colons in R for run-on code execution (try `head(iris); tail(iris)`)

So what? Well, you can exeute each line in turn and check the output. But wait: an object called `.` is not presented in the global environment. No, unless you check 'Show .Last.value in environment listing' in RStudio's settings. Now when you run the line you'll see the '.Last.value' that's been output.

```{r bizarro-example}
iris_bizarre <- iris ->.;
  filter(., Species %in% c("setosa", "versicolor")) ->.;
  group_by(., Species) ->.;
  summarise(., `Mean width` = mean(Sepal.Width)) ->.;
  mutate(., `Mean width` = round(`Mean width`, 1))
```



https://r4ds.had.co.nz/pipes.html


<!-- # I prefer cigars -->

<!-- What was it like in the good old days? -->

<!-- ## Intermediate objects -->

<!-- In the first approach we make a series of intermediate objects, each created using a single function. -->

<!-- ```{r intermediate-objects} -->
<!-- iris_filter <- filter(iris, Species %in% c("setosa", "versicolor")) -->
<!-- iris_group <- group_by(iris_filter, Species) -->
<!-- iris_mean <- summarise(iris_group, `Mean width` = mean(Sepal.Width)) -->
<!-- iris_mutate <- mutate(iris_mean, `Mean width` = round(`Mean width`, 1)) -->

<!-- iris_mutate -->
<!-- ``` -->

<!-- This seems sensible. You can create and interrogate these objects to make sure they do what you want. But this also makes your environment untidy because you have a bunch of halfway-house objects that may serve no standalone purpose beyond being passed into the next object. It could be tricky to keep track of these objects as they swell in your environment, particularly if they get labelled `temp1`, `temp2`, `temp3`, etc.[^past-matt]   -->

<!-- ## Onions -->

<!-- We can do away with all the intermediate steps by nesting the functions inside each other so that functions are applied to the layer below them. -->

<!-- ```{r onion} -->
<!-- iris_onion <- -->
<!--   mutate( -->
<!--     summarise( -->
<!--       group_by( -->
<!--         filter(iris, Species %in% c("setosa", "versicolor")), -->
<!--         Species -->
<!--       ), -->
<!--       `Mean width` = mean(Sepal.Width) -->
<!--     ), -->
<!--     `Mean width` = round(`Mean width`, 1) -->
<!--   ) -->

<!-- iris_onion -->
<!-- ``` -->

<!-- I've called this the onion method. Why? Because it's made up of multiple layers. And it makes you cry a bit to look at it. -->


<!-- # The pipe -->

<!-- The pipe operator, `%>%`, chains function calls together into 'pipelines', which -->

<!-- >semantically changes your code in a way that makes it more intuitive to both read and write -->

<!-- The {magrittr} package brought this paradigm to R and has been adopted by the wildly popular packages of the tidyverse. Pipes have changed fundamentally how people interact with the R programming language.  -->

<!-- # Examples -->

<!-- Let's look at three approaches to some simple data manipulation using (1) intermediate objects, (2) nested functions and (3) pipelines. The first two I'm 'classic approaches'. Let's say I want the mean sepal width of the setosa and versicolor species of iris and round it to one decimal place. -->

<!-- Since this post is about coding style and not 'base R versus the tidyverse', it doesn't matter what functions I use to actually do the data manipulation. I'm using {dplyr} here to make the comparisons easier and because it loads the pipe operator. -->

<!-- ```{r dplyr} -->
<!-- suppressPackageStartupMessages(library(dplyr)) -->
<!-- ``` -->

[^vignette]: Or as [the {magrittr} vignette](https://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html) puts it: "to make your code smokin' (puff puff)"
[^past-matt]: I'm looking at you, past-self.