---
title: "{itdepends} on {lubridate}"
author: Matt Dray
date: '2021-10-24'
slug: lubridate-fns
categories:
  - code
  - tutorial
tags:
  - dependencies
  - itdepends
  - lubridate
  - r
draft: yes
---

```{r setup, include=FALSE}
set.seed(1)
```

# tl;dr

I used [{itdepends}](https://github.com/r-lib/itdepends) to see how CRAN packages depend on [{lubridate}](https://lubridate.tidyverse.org/), which [was not removed from CRAN](What if [the {lubridate} package was removed from CRAN](https://github.com/tidyverse/lubridate/issues/998)?) recently.

# Lubrigate

A test failure in [{lubridate}](https://lubridate.tidyverse.org/) led to hundreds of R developers being emailed about [its potential expulsion from CRAN](https://github.com/tidyverse/lubridate/issues/998), which also threatened the hundreds of packages that depend on it. 

I see the benefit of minimising dependencies. I also understand the drawbacks of reinventing the wheel. Maybe {lubridate} is a good dependency: a simple API, part of the popular tidyverse suite of packages, and it handles stuff you can't be bothered with (like what's 29 February plus one year?).

[Jim Hester spoke at rstudio::conf(2019) about dependencies](https://www.tidyverse.org/blog/2019/05/itdepends/). His [{itdepends}](https://github.com/r-lib/itdepends) package helps you understand the scale and impact of your dependencies.

So, for fun, I'm looking at how {lubridate} is used by packages that import it.

# Get CRAN packages

`CRAN_package_db()` is a convenient function that returns information about packages available on CRAN. We can filter if for the packages that import {lubridate}, i.e. they have {lubridate} in the `Imports` section of their DESCRIPTION file.

```{r cran-db}
library(dplyr, warn.conflicts = FALSE)
library(tidyr)
library(stringr)

cran <- tools::CRAN_package_db()

imports_lubridate <- cran |> 
  filter(str_detect(Imports, "lubridate")) |> 
  pull(Package)

sample(imports_lubridate, 5)  # random sample
```

Right, so that's `r length(imports_lubridate)` packages out of `r nrow(cran)` (`r 100 * round(length(imports_lubridate) / nrow(cran), 2)`%). Is that a lot? Well, the tidyverse package {dplyr}, the Swiss Army knife of data wrangling, is listed in the Imports of `r cran |> filter(str_detect(Imports, "dplyr")) |> nrow()`.

# Install all

So, perhaps this is a little nuts, but we're going to install all the {lubridate}-dependent packages. {itdepends} works with locally-installed packages, you see.

```{r install-packages, warning=FALSE, echo=TRUE, results='hide', message=FALSE, cache=TRUE}
tmp <- tempdir()  # temporary folder

purrr::walk(
  imports_lubridate,
  ~install.packages(
    .x, 
    destdir = tmp, 
    dependencies = FALSE,  # skip installing dependencies
    repos = "https://cran.ma.imperial.ac.uk/"  # mirror
  )
)
```

This takes a little while. There's probably faster methods, like maybe [the {pak} package](https://pak.r-lib.org/), but I just used what worked.

# {itdepends}

{itdepends} is available from GitHub.

```{r install-itdepends, eval=FALSE}
remotes::install_github("jimhester/itdepends")
```

Now we can pass each of these package names to the `dep_usage_package()` function of {itdepends} in a loop. We get back a dataframe for each package, listing each function call it makes and the package that the function comes from.

I've added a mildly unorthodox use of `next`, [borrowed from StackOverflow](https://stackoverflow.com/a/55937737), because I was having trouble with the loop after a failure.

```{r itdepends-loop, warning=FALSE, cache=TRUE}
dep_list <- vector("list", length(imports_lubridate)) |> 
  setNames(imports_lubridate)

for (i in imports_lubridate) {
  
  skip <- FALSE
  
  tryCatch(
    { dep_list[[i]] <- itdepends::dep_usage_pkg(i)
    dep_list[[i]]$focus <- i  },
    error = function(e) { 
      dep_list[[i]] <- data.frame(
        pkg = NA_character_, fun = NA_character_, focus = NA_character_
      )
      skip <<- TRUE 
    }
  )
  
  if (skip) next
  
}
```

I absolutely do not claim this to be the best, most optimised approach. But it works. 

# Counting dependencies

Now that {itdepends} has extracted all the function calls from each of the packages, we can take a look at their frequencies.

## Example

Here's the top 10 most-used functions from the first package alphabetically: {academictwitteR}.

```{r ex-count-fn}
ex_pkg <- "academictwitteR"

dep_list[[ex_pkg]] |> 
  count(pkg, fun, sort = TRUE) |>
  slice(1:5)
```

It's not particularly exciting to know that the top 5 are made up of base R functions like the assignment arrow, the dollar-sign data accessor[^nzd] and the square bracket. We also don't really care about the package's internal functions. Let's filter these out and re-count

```{r ex-count-fn-no-base}
base_pkgs <- c(
  "stats", "graphics", "grDevices",
  "utils", "datasets", "methods", "base"
)

dep_list[[ex_pkg]] |>
  filter(!pkg %in% c(base_pkgs, ex_pkg)) |> 
  count(pkg, fun, sort = TRUE) |> 
  slice(1:10)
```

Aha. We can see immediately that the authors have made use of tidyverse to write their package, since you can see {dplyr}, {tibble}, etc, in there. This makes the use of {lubridate} relatively unsurprising. 

Here's the {lubridate} functions used by this package.

```{r ex-count-lubridate}
dep_list[[ex_pkg]] |>
  filter(pkg == "lubridate") |> 
  count(pkg, fun, sort = TRUE)
```

So this package uses four {lubridate} functions for conversion and formatting of datetimes.

## All packages

Now let's take a look at the function calls across all the packages that import {lubridate}. I'm first going to convert the list of results to a dataframe.

```{r list-to-df}
dep_df <- do.call(rbind, dep_list)
```

### Function use by package

This is a count of the number of uses of each {lubridate} function by each of the the `focus` packages (i.e. the packages we installed).

```{r count-focus-fn-lubridate}
pkg_fn_count <- dep_df |>
  filter(pkg == "lubridate") |>
  count(focus, fun, sort = TRUE)

pkg_fn_count |> slice(1:5)
```

Holy moley, the {PriceIndices} package calls `month()` and `year()`, used to extract elements of a date, over 1400 times.

### Unique function use by package

We can also look at things like the packages that make calls to the greatest number of unique {lubridate} packages. Here's the top 5.

```{r count-unique-lubridate-fns}
fn_distinct_count <- dep_df |>
  filter(pkg == "lubridate") |>
  distinct(focus, fun) |>
  count(focus, sort = TRUE) 

fn_distinct_count |> slice(1:5)
```

So these packages are using more than 10 unique functions from {lubridate}, which is pretty extensive usage.

A quick histogram reveals that a large number of packages are actually using just a single {lubridate} function.

```{r hist-unique-lubridate-fns}
hist(
  fn_distinct_count$n,
  breaks = 30,
  main = "Unique {lubridate} functions used by\npackages importing {lubridate}",
  xlab = "Function count"
)
```

So, of these packages that use just one {lubridate} function, which function is the most frequent?

```{r single-call-pkgs}
focus_one_fn <- fn_distinct_count |>
  filter(n == 1) |>
  pull(focus)

pkg_fn_count |> 
  filter(focus %in% focus_one_fn) |> 
  count(fun, sort = TRUE) |> 
  slice(1:5)
```

Looks like some pretty standard functions, like converting to a date (`as_date()`, `as_datetime()`) or to parse dates with a particular time component (`ymd_hms` for year, month, date, hour, minute, seconds, and `ymd()`).

I think this is interesting: some packages are importing {lubridate} in its entirety to use a single function. And thse functions have base R equivalents with no package-dependency cost. Without diving too deep, this implies that people are using {lubridate} because of syntax familiarity or perhaps because they're already loading other tidyverse packages anyway. 

### Non-unique function use by package

What about total calls?

```{r percent-lubridate}
fn_nondistinct_count <- dep_df |>
  filter(pkg == "lubridate") |>
  count(focus, sort = TRUE)

dep_df |> 
  count(focus) |> 
  left_join(
    fn_nondistinct_count,
    by = "focus",
    suffix = c("_total", "_lub")
  ) |> 
  mutate(percent_lub = round(100 * n_lub / n_total, 1)) |> 
  arrange(desc(percent_lub)) |>
  slice(1:5)
```

And another quick histogram.

```{r nonunique-lubridate-count}
hist(
  fn_nondistinct_count$n,
  breaks = 30,
  main = "Non-unique {lubridate} functions used by\npackages importing {lubridate}",
  xlab = "Function count"
)
```

It'd be interesting to do this for all CRAN packages and their dependencies, but I don't have a personal CRAN mirror and I don't write particularly performant code.

# So

Use whatever dependencies you want, it's up to you. I write joke packages that I don't put on CRAN, so.

[^nzd]: You should [be aware of the international conspiracy](https://www.rostrum.blog/2021/07/15/dollar-dollar/) behind the use of this symbol in R.