---
title: "Reveal a hidden gorilla with {magick}"
author: Matt Dray
date: '2021-10-05'
slug: gorilla
categories:
  - code
  - tutorial
tags:
  - magick
  - r
  - raster
---

# tl;dr

You can convert a line drawing to datapoints with [a sprinkle of {magick}](https://docs.ropensci.org/magick/).

# Ape escape

Have you seen [that video](http://www.theinvisiblegorilla.com/gorilla_experiment.html) where you're so focused on counting basketball passes that you fail to see the gorilla moving across the screen?

This kind of selective attention was studied by two researchers, Yanai and Lercher, who provided subjects with a dataset that looked like a gorilla when plotted. [The gorilla was found less often if the subjects were also given a hypothesis to investigate](https://www.biorxiv.org/content/10.1101/2020.07.30.228916v1).

The study got some attention on Twitter last week. As a result, [Isabella Velásquez wrote a great blogpost](https://ivelasq.rbind.io/blog/reticulate-data-recreation/) where she recreated the dataset using R and Python in tandem via [the {reticulate} package](https://rstudio.github.io/reticulate/).

I had a go at creating the dataset with base R and [the excellent {magick} package for image manipulation](https://docs.ropensci.org/magick/).

# Point it out

The jpeg image file used in the original paper can be downloaded from classroomclipart.com to a temporary location on your machine.

```{r download-file}
download.file(
  paste0(
    "https://classroomclipart.com/images/gallery/",
    "Clipart/Black_and_White_Clipart/Animals/",
    "gorilla-waving-cartoon-black-white-outline-clipart-914.jpg" 
  ),
  tempfile(fileext = ".jpg")
)
```

We can read the file into R with {magick}.

```{r orig-img, fig.alt="A line drawing of a cartoon gorilla waving."}
img <- 
  list.files(tempdir(), pattern = ".jpg$", full.names = TRUE) |>
  magick::image_read()

img
```

With other {magick} functions we can:

* reduce to two distinct colours only (i.e. for the lines and background), which makes it easier to filter the data later
* convert from an image to point data

```{r wrangle-magick}
go <- img |>
  magick::image_quantize(2) |>  # colour reduction
  magick::image_raster() |>     # as x-y data
  as.data.frame()

head(go)
```

And to prove we only have two colours (off-white for the background, grey for the lines):

```{r colours}
unique(go$col)
```

Now we can:

* reverse the order of the `y` values so the gorilla faces the same way as in the paper
* filter to retain only the datapoints that represent lines
* rescale the `x` and `y` to create 'Body Mass Index' (BMI)[^bmi] and 'steps' variables

```{r wrangle-base}
go$y     <- rev(go$y)
go       <- go[go$col != "#fefefeff", ]
go$bmi   <- go$y / max(go $y) * 17 + 15
go$steps <- 15000 - go$x * 15000 / max(go$x)

head(go)
```

You may have noticed that the image has a watermark. We could have removed it earlier with {magick}, but can do it now by filtering out the datapoints in that corner.

```{r rm-logo}
go$logo <- ifelse(go$bmi < 16 & go$steps < 5500, TRUE, FALSE)
go      <- go[!go$logo, ]
```

This leaves us with `r nrow(go)` datapoints. We can follow the original study by taking a sample and splitting the results into 'female' and 'male' groups, weighted so that the female group has higher step counts.

```{r group-datasets}
go_smp       <- go[sample(nrow(go), 1768), ]
go_smp$rnorm <- rnorm(nrow(go_smp), mean = 0, sd = 10)
go_smp$index <- go_smp$steps * (1 + go_smp$rnorm)
go_smp$group <- 
  ifelse(go_smp$index < median(go_smp$steps), "F", "M") |>
  as.factor()

head(go_smp[, c("bmi", "steps", "group")])
```

Now finally to plot the datasets side-by-side.

```{r two-plots, fig.alt="Two side-by-side plots of steps (x) against BMI (y) where both sets of datapoints look like a cartoon gorilla waving."}
par(mfrow = c(1, 2))

with(
  go_smp[go_smp$group == "F", ],
  plot(
    steps, bmi, xlim = c(0, 15000),
    pch = 16, cex = 0.5, col = "blue",
    xlab = "Steps", ylab = "BMI", 
  )
)

with(
  go_smp[go_smp$group == "M", ],
  plot(
    steps, bmi, xlim = c(0, 15000),
    pch = 16, cex = 0.5, col = "red",
    xlab = "Steps", ylab = "BMI"
  )
)
```

I see them!

This has been a bit overengineered and could be generalised, but it gives a gist of how you might go about converting an image to a dataframe of x and y positions.

At worst, this is a reminder not to trust researchers and to always check for unexpected gorillas.

---
<details><summary>Session info</summary>
```{r sessioninfo, echo=FALSE}
sessioninfo::session_info()
```
</details>

[^bmi]: Check out [a recent episode of the _Maintenance Phase_ podcast](http://maintenancephase.com/) (dated 2021-08-03) about the troublesome history and development of BMI as a metric.
