---
title: 'Advanced R: anchovy pizza'
author: 'Matt Dray'
date: '2021-01-23'
slug: adv-r-names
  - code
tags:
  - advanced-r
  - book
  - food
  - learning
  - r
draft: yes
---

# tl;dr

R is quirky. I bought [Hadley Wickham](http://hadley.nz/)'s [Advanced R](https://adv-r.hadley.nz/) book[^buy] to help me better understand it. I plan to write about some thought technologies as I go.

Today: names and values (chapter 2) as pizza recipes.[^royale]

# Binding basics

Welcome to the pizzeria. We store our toppings as character-vector objects.

```{r}
pizza_napoletana <- c("mozzarella", "tomato", "anchovy")
```

The English version of the menu calls it 'Neapolitan pizza', but it's exactly the same thing.

```{r}
neapolitan_pizza <- pizza_napoletana
```

To reflect this, our recipe book uses a unique code for each toppings object and we can bind as many names to it as we want.

In other words, the pizzaiolos don't care whether you asked for a 'pizza alla napoletana' or a 'Neapolitan pizza'. The recipe is exactly the same one.

In our pizzeria we store the recipes with a special code. Both pizza names point to the same underlying recipe.

```{r}
lobstr::obj_addr(pizza_napoletana)
lobstr::obj_addr(neapolitan_pizza)
```

<div class="tip"> `r emo::ji("key")` <b>Key point</b>
<br>
The object is real. The name points to it.
</div>

# Copy on modify

Pizza from Puglia is being introduced to the restaurant. It has the same basic toppings as the one from Napoli, but it has onion instead of anchovies.

We can just copy the Neapolitan recipe and make the change.

```{r}
pizza_pugliese <- pizza_napoletana
pizza_pugliese[[3]] <- "onion"
```

Even though we copied the mozzarella and tomato from another recipe, the switch to onion means we actually have an entirely new recipe that warrants an entirely new code.

```{r}
lobstr::obj_addr(pizza_napoletana)
lobstr::obj_addr(pizza_pugliese)
```
Here's how the menu might look for these pizzas.[^damichele] Same underlying object for the first two regardless of the name, while the third is entirely different even though it was made by copying the first.

```{r}
tibble::tribble(
  ~language, ~pizza, ~code,
  "ITA", "Pizza alla napoletana", lobstr::obj_addr(pizza_napoletana), 
  "ENG", "Neapolitan pizza", lobstr::obj_addr(neapolitan_pizza), 
  "ITA", "Pizza pugliese", lobstr::obj_addr(pizza_pugliese)
)
```

# Lists

This knowledge management system is a bit inefficient: the 'mozzarella' and 'tomato' toppings  exist twice in our recipe book, once for each pizza.

We could store our recipes as lists instead so that each unique topping is stored once and each recipe refers to that one instance.

```{r}
pizza_napoletana <- list("mozzarella", "tomato", "anchovy")
pizza_pugliese <- pizza_napoletana
pizza_pugliese[[3]] <- "onion"
```

In our recipe book, the pizza alla napoletana (`1` in the diagram below) and pizza pugliese (`5`) recipes now make the same reference to mozzarella (`2`) and tomatoes (`3`), as well as separate references to their additional toppings of anchovies (`4`) and onions (`6`).

```{r}
lobstr::ref(pizza_napoletana, pizza_pugliese)
```

# Data frames

Let's say the waiter is taking orders. A couple have ordered a napoletana and a pugliese. The waiter stores these as columns in a `data.frame`.

```{r}
order <- data.frame(
  napoletana = c("mozzarella", "tomato", "anchovy"),
  pugliese = c("mozzarella", "tomato", "onion")
)

order
```

The customer who ordered the napoletana changed their mind — more anchovies!

```{r}
order_update <- order
order_update[3, 1] <- "anchovies (extra)" 
order_update
```

The waiter copies the napoletana order and changes the `anchovies` to `anchovies (extra)`. 

Since a data.frame is a list of vectors, the napoletana vector (`2` in the diagram below) in the initial order (`1`) is modified on copy (`5`) in the updated order (`2`). The pugliese order (`3`) didn't change between orders.


```{r}
lobstr::ref(order, order_update)
```

The other customer is jealous and chooses to add extra cheese to their order.

```{r}
order_final <- order_update
order_final[1, 2] <- "mozzarella (extra)"
order_final
```

The napoletana (`2`) and pugliese (`3`) vectors in the original order (`1`) have both changed (`5`, `6`) in the final order (`4`).

```{r}
lobstr::ref(order, order_final)
```
A data.frame holds pointers to vectors. The initial order and the updated order are different objects that point to the same pugliese vector, but different napoletana vectors.

The final order is a different data.frame object again, and it points to two entirely different napoletana and pugliese vectors compared to the initial order.

---
<details><summary>Session info</summary>
```{r sessioninfo, echo=FALSE}
sessioninfo::session_info()
```
</details>

[^buy]: You can [buy the book](https://www.amazon.com/gp/product/0815384572/ref=as_li_tl?ie=UTF8&tag=hadlwick-20&camp=1789&creative=9325&linkCode=as2&creativeASIN=0815384572&linkId=9f8b9df5da858c12a5e9905aea66bf08), or [view it for free online](https://adv-r.hadley.nz/).
[^royale]: Initially I went with a metaphor about how a Quarter Pounder with Cheese is called a Royale with Cheese in Paris (or indeed, a Krusty Burger with Cheese is called a Quarter Pounder with Cheese at McDonald's) despite them being the same thing, but the pop culture reference was better than the actual utility of the metaphor. 
[^damichele]: Of course, if you're really serious about pizza, you only offer marianara and margherita [as per L'antica Pizzeria Da Michele](http://www.damichele.net/le-nostre-pizze/), which is where I took the photos that are in this post.