---
title: "Basic base R dataframes"
author: 'Matt Dray'
date: '2022-08-20'
slug: base-df
categories:
  - code
  - tutorial
tags:
  - base
  - r
draft: yes
---

# tl;dr 

Placeholder.

# Intro

Jump to:

* Example dataframe
* Basic summaries
* Access data
* Alter a value
* Rename columns
* Add new columns
* Subset data
* Rename factor levels
* Reorder factor levels
* Collapse factor levels
* Append dataframes
* Merge dataframes

# Example dataframe

We are going to build a dataframe by specifying three vectors as objects using the gets notation (<-), and then arranging them into columns using the data.frame function.
This dataframe is used as the starting point for the manipulations covered in this post.

```{r}
VAR1 <- c("Ecol.Lett", "Annu.Rev.Ecol.Evol.S", "Ecol.Monogr", "Front.Ecol.Environ","Glob.Ecol.Biogeogr", "Glob.Change.Biol", "Ecology", "J.Ecol") # Journal titles
VAR2 <- c("BP", "ARI", "ESA", "ESA", "BP", "BP", "ESA", "BP") # Abbreviated publisher names
VAR3 <- c(7.995, 7.019, 5.21, 3.644, 3.477, 3.425, 3.246, 3.145) # SCImago Journal Rank (SJR)
data <- data.frame(VAR1, VAR2, VAR3)
data
```

# Basic summaries

The following functions allow you to look at the dataframe, or parts of it, and some simple summary statistics.

```{r}
data # Show full dataframe
```



```{r}
head(data, n = 4) # Show first four rows, default is six
```


```{r}
names(data) # List variable (column) names
```


```{r}
nrow(data) # Number of rows
```


```{r}
ncol(data) # Number of columns
```


```{r}
summary(data) # Factor levels frequencies, quartiles and integer ranges
```


```{r}
str(data) # STRucture of dataframe (data type, number of levels, first few values)
```

# Access data

You will often need to access individual data points or groups of data points.
This is done using dollar ($) or square bracket ([]) notation, or a hybrid of the two, to call upon specific rows and columns
```{r}
data$VAR2 # Dollar notation; returns values from column 'VAR1' in dataframe 'data'
```


```{r}
data[2] # Square-bracket notation (indexing); returns column 1 in dataframe 'data'
```

```{r}
data[3, 2] # Value in 3rd row, 2nd column
```

```{r}
data$VAR2[3] # Hybrid notation; returns value from the third row of column 'VAR1' in dataframe 'data'
```

# Alter a value

Once you have located the value(s) you want changed, you can use the gets notation (<-) to specify that location as a new value.

```{r}
data$VAR3[1] # Shows incorrect value of 7.995
```

```{r}
data$VAR3[1] <- 7.996 # You can use $, [], or a hybrid notation to locate data
data$VAR3[1] # Value corrected to 7.996
```

# Rename columns

Informative column headers can be used to replace boring ones from your dataframe (e.g. VAR1 in our example).

The names function allows us to look at the column names, which we can change using the gets notation (<-).

You can change the column headers one by one, but it is easier to concatenate (c) a list of new column headers if you want to change several at once.

```{r}
names(data) <- c("Title", "Publisher", "SJR") # Rename multiple columns
names(data)[3] <- "SJR" # Rename the third column only
data
```

# Add new columns

This is similar to the in the Example dataframe section above, where you specify a concatenated (c) list, or vector, of values to make up the new column.
Create a new empty column in the dataframe using dollar notation (e.g. dframe1$column.one creates a new column called column.one in a dataframe called dframe1) and fill it with your vector using the gets notation (<-)

```{r}
data$H.index <- c(127, 125, 98, 57, 71, 126, 172, 102) # Add new column called 'H.index' with listed data
data
```

# Subset data

Subsetting allows you to focus on certain sections of your dataframe (e.g. you have a dataframe about Pokémon, but you want to narrow your analysis down to bug-types only).
You can select the data you want by matching it to a list, using using logical operators to simplify the process: == means equal to; != means not equal to; %in% means match to a list of options.

(1) Use the 'subset' function

```{r}
data.BP <- subset(data, Publisher == "BP") # Dataframe containing publisher BP only
data.notBP <- subset(data, Publisher != "BP") # Dataframe containing all publishers except BP
data.ARIandESA <- subset(data, Publisher %in% c("ARI", "ESA")) # Dataframe containing publishers ARI and ESA only
```

(2) Use square-bracket notation (indexing)

```{r}
data.BP <- data[data$Publisher == "BP", ] # Dataframe containing publisher BP only
data.notBP <- data[data$Publisher != "BP", ] # Dataframe containing all publishers except BP
data.ARIandESA <- data[data$Publisher %in% c("ARI", "ESA"), ] # Blank space after the comma in the square brackets indicates inclusion of all columns
data.ARIandESA <- data[data$Treatment %in% c("ARI", "ESA"), c(2, 3)] # If we only want some of the columns (e.g. columns 2 and 3)
```

# Rename factor levels

This is similar to renaming columns (above), but uses the levels function instead of the names function.

```{r}
summary(data$Title) # List original level names
```

```{r}
levels(data$Title) <- c("AREES", "EL", "EM", "Ecol", "FEE", "GCB", "GEB", "JE") # Replacing level names as ordered in the summary (alphabetically)
summary(data$Title) # List new level names
```

# Reorder factor levels

By default, factors are listed alphabetically, but you may want them to appear in a different order (this allows you to plot these data in the order you specify, too).
```{r}
summary(data$Title) # Check the levels
```


```{r}
data$Title <- factor(data$Title, levels = c("EM", "AREES", "FEE", "GEB", "EL",
"Ecol", "GCB", "JE")) # Order the levels as you please
summary(data$Title) # Check order
```

# Collapse factor levels

This can be done to simplify a model; perhaps there is not enough of a difference between Factors A and B, so they can be considered together.

(1) Create new levels to collapse current levels into

```{r}
data$Collapsed.titles <- data$Title # Copy 'Title' column to new column 'Collapsed.title'
levels(data$Collapsed.titles) # Factor list is AREES, EL, EM, Ecol, FEE, GCB', GEB, JE
```

```{r}
levels(data$Collapsed.titles) <- c(levels(data$Collapsed.titles), "Top4", "Next4") # Introduce levels to collapse others into
```

(2) Collapse the current levels into the new levels

```{r}
data$Collapsed.titles[data$Collapsed.titles %in% c("EL", "AREES", "EM", "FEE")] <- "Top4" # Four journals with highest SJR
data$Collapsed.titles[data$Collapsed.titles %in% c("GEB", "GCB", "Ecol", "JE")] <- "Next4" # Four journals with next highest SJR
```

(3) Remove the now-unwanted old levels

```{r}
summary(data$Collapsed.titles) # Unwanted levels still present
```

```{r}
data <- droplevels(data) # Drop the unwanted levels
summary(data$Collapsed.titles) # Unwanted levels gone
```

# Append dataframes

This allows you to attach two dataframes together vertically ('one on top of the other').
The column names in each dataframe must match for this to work.

(1) Set-up two example dataframes to append

```{r}
Title <- c("EL", "AREES", "EM", "FEE", "GEB", "GCB", "Ecol", "JE")
Publisher <- c("BP", "ARI", "ESA", "ESA", "BP", "BP", "ESA", "BP")
SJR <- c(7.996, 7.019, 5.21, 3.644, 3.477, 3.425, 3.246, 3.145)
H.index <- c(127, 125, 98, 57, 71, 126, 172, 102)
data <- data.frame(Title, Publisher, SJR, H.index)
data
```

```{r}
Title <- c("Ecog", "DD")
Publisher <- c("BP", "BP")
SJR <- c(2.87, 2.811)
H.index <- c(72, 60)
Published.2012 <- c(134, 116)
data2 <- data.frame(Title, Publisher, SJR, H.index, Published.2012)
data2
```

(2) Check column names match

```{r}
names(data) == names(data2) # Check that columns match ('data2' has an extra column)
```

```{r}
data2$Published.2012 <- NULL # Effectively deletes the column
names(data) == names(data2) # Check again that columns match
```

(3) Append

```{r}
data.2and3 <- rbind(data, data2) # Append the dataframes
data.2and3 # Check for success
```

# Merge dataframes

You have a second dataframe with further information about levels in the first dataframe.
For instance, our example dataframe can be merged with a dataframe containing further information about the publishers.

(1) Set-up two example dataframes to merge

```{r}
Title <- c("EL", "AREES", "EM", "FEE", "GEB", "GCB", "Ecol", "JE")
Publisher <- c("BP", "ARI", "ESA", "ESA", "BP", "BP", "ESA", "BP")
SJR <- c(7.996, 7.019, 5.21, 3.644, 3.477, 3.425, 3.246, 3.145)
H.index <- c(127, 125, 98, 57, 71, 126, 172, 102)
data <- data.frame(Title, Publisher, SJR, H.index)
data
```

```{r}
Publisher <- c("BP", "ARI", "ESA")
Country <- c("UK", "USA", "USA")
data3 <- data.frame(Publisher, Country) # Note that this dataframe contains further information about factors in the dataframe 'data' (the country that each publisher operates from)
data3
```

(2) Merge the two dataframes

```{r}
data.merge <- merge(data, data3, by = "Publisher") # We've now added extra information
data.merge
```



# Todo

transform() and within() to add columns
reshape() to pivot
sort()

---
<details><summary>Session info</summary>
```{r sessioninfo, echo=FALSE}
sessioninfo::session_info()
```
</details>
