---
title: 'One weird trick to {monetize} your R package'
author: Matt Dray
date: 2023-08-01
slug: monetize
categories:
  - code
  - package
tags:
  - monetize
  - r
draft: yes
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(eval = FALSE)  # depends on local storage, so hard-coded
library(monetize)
```

![](){}

# tl;dr

The {monetize} package gives you inspiration for monetising your R package. Developers rise up!

# Free as in free

There's one good reason why SAS, SPSS and Stata[^r] are such successful and beloved statistical tools: money.

For come reason, R remains free and open source. This is despite the rumours[^first] that Ihaka & Gentleman originally wanted users to be charged a literal (New Zealand) dollar[^nz] every time they use the `$` symbol to access an element from an object.

Us package developers should seize the initiative and harvest the sweet, sweet bounty of our labour. But how? I think I can help you.

# Free as in not quite

Introducing the {monetize} R package! Serious entrepreneurs can install from GitHub[^dont].

```{r install}
install.packages("remotes")  # if not yet installed
remotes::install_github("matt-dray/monetize")
library(monetize)
```
```
Welcome to the monetize(TM) package FREE TRIAL version!
Try our EXCLUSIVE add_one() function!
Try watch_ad() to gain MEGA COINS and increase your MEMBER LEVEL for REWARDS!
And now a quick message from our sponsor:
  🧃 IMBIBE ACME(TM)-BRAND FRUIT PULP (7% FEWER INSECT COMPONENTS)! 🪳
```

The package contains the exclusive `add_one()` function, which surprisingly doesn't come pre-installed with base R. A gap in the market!

Users should speculate to accumulate, so there's a small cost to use the function. In {monetize}, each use of a function costs 1 MEGA COIN.

Because I am so kind, I have chosen for users to receive a 30-day FREE TRIAL of the package and to receive 3 MEGA COINS as an exclusive NEW-MEMBER PERK.

One use of the function costs the low, low price of 1 MEGA COIN.

```{r add-one}
add_one(1)
```
```
📅 You have 30 days left of your FREE TRIAL!
🏅 Your MEMBER LEVEL is 1! Try watch_ad()!
💰 Your MEGA COIN balance is now 2! Try watch_ad()!
[1] 2
```

Along with the result, the function helpfully prints the number of days left in your FREE TRIAL, your MEMBER LEVEL and the number of MEGA COINS you have left.

What's a MEMBER LEVEL? Well, at MEMBER LEVEL 1, the function can't guarantee a correct result. 20% of the time there'll be an off-by-one error. But R users index from 1 anyway, so they'll be used to that.

```{r add-one-wrong}
add_one(1)
```
```
📅 You have 30 days left of your FREE TRIAL!
🏅 Your MEMBER LEVEL is 1! Try watch_ad()!
💰 Your MEGA COIN balance is now 1! Try watch_ad()!
[1] 3
```

How can you get the correct result 100% of the time? That's an exclusive for MEMBER LEVEL 2. How do you reach MEMBER LEVEL 2? What happens if you run out of MEGA COINS?

```{r no-coins}
add_one(1)
```
```
📅 You have 30 days left of your FREE TRIAL!
🏅 Your MEMBER LEVEL is 1! Try watch_ad()!
Error: 😭 You'll need more MEGA COINS to re-use this function! Try watch_ad()!
```

Answer: simply watch some ads! I've managed to get an ad deal with the famous [Acme(TM) Corporation](https://en.wikipedia.org/wiki/Acme_Corporation). Maybe you could get a sponsorship deal from a third-party whose interests align with your users' interests; maybe SAS, SPSS or Stata would be interested?

Anyway, the user can simply `watch_ad()` in exchange for MEGA COINS: I've allowed the user to employ their FREE WILL to watch a `"short"` ad for 1 MEGA COIN or a `"long"` ad for 3 MEGA COINS.

```{r watch-add}
watch_ad("short")
```
```
🌭 CONSUME ACME(TM)-BRAND RECONSTITUTED MEAT-LIKE CYLINDERS! 🐕
🌭 CONSUME ACME(TM)-BRAND RECONSTITUTED MEAT-LIKE CYLINDERS! 🐕
🌭 CONSUME ACME(TM)-BRAND RECONSTITUTED MEAT-LIKE CYLINDERS! 🐕
🌭 CONSUME ACME(TM)-BRAND RECONSTITUTED MEAT-LIKE CYLINDERS! 🐕
🌭 CONSUME ACME(TM)-BRAND RECONSTITUTED MEAT-LIKE CYLINDERS! 🐕
Congratulations! Your new MEGA COIN total is 1!
```

Reach MEMBER LEVEL 2 after gaining 10 all-time MEGA COINS and MEMBER LEVEL 3 after 20 MEGA COINS.

What's the exclusive MEMBER LEVEL 3? Of course, the answer will be correct... and comes with a randomised FREE HAT!

```{r hat}
add_one(1)
```
```
📅 You have 30 days left of your FREE TRIAL!
🏅 Your MEMBER LEVEL is 3! Try watch_ad()!
💰 Your MEGA COIN balance is now 19! Try watch_ad()!
🎩 
 2 
```

Watch the cash roll in.

# Free as in absolutely not

I hope this has been an inspirational eye-opener for you and that one day you can be as rich as me. Oh, you didn't know I was rich? How else could I afford to keep this absolute juggernaut of a blog going?

To demonstrate how little I personally need them, here are some more more ideas. You could try:

* releasing version updates as paid addons, like videogame DLC
* free access to only the first page of your package documentation
* lootboxes to allow users to gamble—I mean, test their precognitive skills—for new functionality (`add_two()` could be a common drop, while `multiply_by_two()` is ultra rare)

You are welcome.

---
<details><summary>Session info</summary>
```{r eval=TRUE, sessioninfo, echo=FALSE}
sessioninfo::session_info()
```
</details>

[^r]: Curious why they all start with 'S'? It's in deference to 'R', which comes first in the alphabet and in being really cool.
[^dont]: Don't.
[^ruserdir]: I consider this my hattrick of posts on `R_user_dir()`. I wrote about it as a way to [cache data from the PokéAPI in {trapinch}](https://rostrum.blog2023/02/02/trapinch) and [store cyber pet blueprints in {tamRgo}](https://rostrum.blog/2022/11/13/tamrgo), the latter of which led to a post on [saving high-score data with {hiscore}](https://rostrum.blog/2023/07/15/hiscore).
[^first]: You heard it here first... because it was me that made it up.
[^nz] Did you also know that the use of the New Zealand dollar is part of [a wider conspiracy](htpps://rostrum.blog/2021/07/15/dollar-dollar)? The sheep dip goes deep on this one.