---
title: '{gert} going with Git'
author: Matt Dray
date: '2021-05-05'
slug: gert
categories:
  - code
  - tutorial
tags:
  - gert
  - git
  - github
  - r
draft: yes
---

```{r, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
library(gert)
```

# tl;dr





Learning Git on the command line is an extra burden for users switching from point-and-click workflows to R-led reproducible and version-controlled workflows. The {gert} package uses R functions and objects to interact with Git instead, which might make version control easier to use consistently for these users.


# Different version control versions

I try to help people transition their analysis from point-and-click workflows to reproducible and version-controlled code-based projects. 

I think it's relatively easy to sell the idea of version control. It's harder, in my experience, to make the process stick. I think a big part of that is how daunting it can be to learn R and then also have to learn Git and the command line.

If you're already learning R, then why not use an R interface to Git?

I'm advocating for a code-led version control process that's . It would be strange to pry people away from point-and-click workflows and then use the point-and-click GUI available in RStudio, for example, though I can see how such tools minimise friction versus writing on the command line.

# Summary

There's a lot of git commands and a lot of {gert} funcitons that replicate them. For purposes of this post, and the focus on simple add-commit-push Git workflows, there's a few main commands to consider:

| Command | Git | {gert} |
| :---- | :---- | :---- |
| Initiate Git | `git init` | `git_init("<url>")` |
| Clone a repo | `git clone <url>` | `git_clone("<url>")` |
| Pull changes | `git pull` | `git_pull()` |
| Check for changes | `git status` | `git_status()` |
| Stage changes | `git add <file>` | `git_add()` |
| Commit | `git commit -m "<message>"` | `git_commit("<message>")` |
| Push | `git push` | `git_push()` |
| Check current branch | `git branch` | `gert::git_branch()` |
| Create branch | `git branch -b <branch>` | `git_branch_create("<branch>")` |
| Switch to branch | `git checkout <branch>` | `git_branch_checkout("<branch>")` |

# Approaches

This post also isn't about setting up Git and GitHub. I recommend you check out the [Happy Git and GitHub for the UseR](https://happygitwithr.com/) book by Jenny Bryan, the STAT545 TAs and Jim Hester. I've also written an earlier post about Git and GitHub that explains [the one-time setup tasks](https://www.rostrum.blog/2019/10/21/git-github/#one-time) you'll need to do.

## Command-line workflow

Remember, these are being typed directly into the command line in an interactive session. I've used the shell prompt `$` as a reminder.

```
$ git init
```
```
```

```
$ git remote add("<url>")
```
```
```

```
$ git status
```
```
```

```
$ git add "<file>"
```
```
```

```
$ git commit -m "<message>"
```
```
```

```
$ git push origin main
```
```
```

## {gert} workflow

Instead of opening a terminal to write git commands, {gert} lets you write functions in the R console.

This is good for beginners or people not confident with the command line.

This means you can write R functions

You're returned familiar R objects with the Git information, which you can further interrogate with R code. You don't have to understand how to write at the command line to explore the information.

{gert} doesn't have functions to replicate all Git commands, but that's not really a problem for the simpler status-add-commit-push workflow.

Remember, these are being typed directly into the R console in an interactive session. I've used the R prompt `>` as a reminder.

```{r}
> install.packages("gert")
> library(gert)
```
```{r}
# Linking to libgit2 v1.1.0, ssh support: YES
# Global config: /Users/matt/.gitconfig
# Default user: Matt Dray <email>
```

After attaching the package, you get some information: SSH support is active, the filepath of the global git config file and the name and email address of the user to whom any commits will be ascribed.

The function names pretty much replicate Git commands, so `git status` is `git_status()`. There's some handy additional funcitons like `git_commit_all("<message>")` that automatically stages and commits all changes, i.e. it runs `git add .` and `git commit -m <message>` in one go.

```{r}
> git_init()
```

```{r}
> git_remote_add("git@github.com:matt-dray/test.git")
```

```{r}
> usethis::use_git_ignore(c(".Rproj.user", ".DS_Store"))
```
```{r}
# ✓ Adding '.Rproj.user', '.DS_Store' to '.gitignore'
```


```{r}
> git_status()
```
```{r}
# # A tibble: 2 x 3
#   file       status staged
#   <chr>      <chr>  <lgl> 
# 1 .gitignore new    FALSE 
# 2 test.Rproj new    FALSE 
```

```{r}
> git_add(c(".gitignore", "test.Rproj"))
```
```{r}
# A tibble: 2 x 3
#   file       status staged
#   <chr>      <chr>  <lgl> 
# 1 .gitignore new    TRUE  
# 2 test.Rproj new    TRUE 
```

```{r}
> git_commit("Add gitignore")
```
```{r}
# [1] "38608b0fe0cd7258919d05acd94934013ad04333"
```

```{r}
> git_push()
```
```{r}
# Trying to authenticate 'git' using ssh-agent...
# Trying to authenticate 'git' using provided ssh-key...
# Transferred 4 of 4 objects...done!
# [status] refs/heads/master: unchanged
# [new]     38608b0fe0cd7258919d refs/remotes/origin/master
```

# Bonus

## Git prompt

Sometimes you forget what branch you're on or if any changes have been made since the last commit. To prevent you from checking `git status` continually, you can  customise your R prompt like this, where the asterisk indicates a change:

```
main* >
```

[The {prompt} package](https://github.com/gaborcsardi/prompt) has functions to do this. You can run {prompt} code on startup by adding it to your RProfile file. The {usethis} package will help you find and open your RProfile if you don't know where it is.

```{r}
install.packages(c("prompt", "usethis"))
usethis::edit_r_profile()
```

Your RProfile file will open. Add the following:

```{r}
if (interactive()) prompt::set_prompt(prompt::prompt_git())
```

After refreshing, you'll see the prompt has changed. This will happen for every R instance you open from now on. You can `prompt::suspend()` to return to the classic prompt, but you can `prompt::restore()` it at any time.

You can set even more informative prompts, like `prompt::prompt_fancy()`, or create your own, as per [Matt Kerlogue's walkthrough](https://lapsedgeographer.london/2020-11/custom-r-prompt/).